\documentclass[12pt]{article}

% -------------------------------------------------------------------
% Packages
% -------------------------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{geometry}
\usepackage{hyperref}
\geometry{margin=1in}

\title{Spherepop Calculus and Amplitwistor Computation:\\
A Geometric--Thermodynamic Foundation for AI}
\author{Flyxion}
\date{\today}

\begin{document}
\maketitle

% -------------------------------------------------------------------
% Abstract
% -------------------------------------------------------------------
\begin{abstract}
We introduce \emph{Spherepop Calculus} (SPC), a foundational
computational formalism in which the primitives of computation are
\emph{geometric collapse}, \emph{merge}, and the dynamics of a
background plenum endowed with scalar, vector, and entropy fields.
Computation is interpreted as geometric evolution, not symbolic
rewriting. We show that classical paradigms including lambda calculus,
functional programming, logical inference, neural architectures, and
tensor contraction formalisms arise as \emph{coordinate projections}
or \emph{degenerations} of SPC, rather than independent foundations.

We then develop a complex--analytic representation of SPC via
\emph{amplitwistors}, in which collapse corresponds to holomorphic
scattering, merge to contour gluing, and global computation to
amplitwistor integration constrained by thermodynamic flow. The result
is a unified computational substrate that is geometric, thermodynamic,
holomorphic, and inherently nonlocal. SPC thus provides a principled
foundation for AI that subsumes and geometrically explains neural,
logical, and tensor--based reasoning systems.
\end{abstract}

% -------------------------------------------------------------------
% Introduction
% -------------------------------------------------------------------
\section*{Introduction}

Foundations of computation have historically relied on symbolic
formalisms such as lambda calculus and Turing machines, later
complemented by vector--based and tensor--based methods in machine
learning. These approaches introduce computational primitives (terms,
weights, embeddings) that are external to the dynamics they describe.
In each case, computation is implemented by \emph{syntactic} or
\emph{algebraic} manipulation, rather than derived from geometric or
physical principles.

This work begins instead from geometry. We propose \emph{Spherepop
Calculus} (SPC), where the fundamental operations are \emph{collapse}
(``pop'') and \emph{merge} of geometric regions (``spheres'') in a
dynamical plenum. Semantics is intrinsic: a term denotes a geometric
configuration, and computation corresponds to geometric evolution
driven by thermodynamic constraints.

Our first objective is to show that classical computational paradigms
arise as restricted viewpoints on SPC. Neural and tensor--based
systems are obtained by local linearization and flattening of boundary
curvature. Logical deduction appears as thresholded collapse. Lambda
calculus and functional programming are recovered when geometric
semantics degenerate to purely syntactic abstraction and application.

Our second objective is to lift computation into complex geometry
using \emph{amplitwistors}, generalizing twistor methods in
mathematical physics.\footnote{For accessible geometric insight into
twistor constructions and complex structure, see T.~Needham's
expository work on complex geometry \cite{Needham}.} Under this
representation, collapse becomes holomorphic scattering; merge becomes
composite interaction; computation becomes contour integration in
amplitwistor space; and inference emerges as thermodynamic
modification of holomorphic structure rather than symbolic deduction
or tensor contraction.

The paper proceeds by defining SPC, establishing its operational and
geometric semantics, demonstrating confluence and computational
adequacy, and then embedding functional and tensor paradigms as
degenerate projections. The amplitwistor formulation is then developed
to yield a geometric--thermodynamic foundation for AI.

\vspace{1em}


\section{Motivation}

Modern AI formalisms rely on representational scaffolding.
Tensors communicate by contractive products along arbitrary index
sets; graphs are constructed to express relational structure between
symbols; and logical systems manipulate terms without spatial
semantics. Each of these formalisms deliberately avoids,
or postpones, geometry. Yet their effectiveness in deep learning
reveals a latent geometric content: internal vector representations
behave as though they encode angles, distances, and flows, but the
geometry remains implicit and ungoverned.

This implicit geometry manifests in awkward inductive biases.
Neural networks must be regularized to approximate locality; graph
neural networks impose relational structure externally; attention
mechanisms reconstruct spatial association by means of dot products
in a latent embedding space. These constructions achieve geometric
effects by indirect means. They rely on explicit scaffolding because
there is no native geometric primitive.

Spherepop calculus eliminates the scaffolding by placing geometry at
the basis of computation. Spheres, pops, and flows are the
fundamental operators, and all subsequent computational phenomena
are derived from these primitives. No indexing convention, no
synthetic tensor product, and no extrinsic graph is required.


\section{Spherepop as Primitive Geometry}

Let $P$ denote a plenum manifold supporting a scalar potential
$\Phi$, vector flow $v$, and entropy density $S$.\footnote{Derived
from the RSVP triad.} A \emph{sphere} is a compact submanifold
$S\subset P$ with boundary $\partial S$ and embedding
$i_S:S\hookrightarrow P$. A \emph{pop} is a boundary-respecting
morphism $\pi:(S,\partial S)\to (P,i_S(S))$ whose differential
$D\pi$ induces a first-order flow $v_\pi$. Intuitively, a pop
collapses the internal configuration of $S$ and radiates geometric
effects outward along the boundary.

We emphasize that spheres and pops are not metaphors, but primitive
constructors. The existence of nested spheres corresponds to
hierarchical scopes; their interaction and dissolution constitute
computation. Classical computational constructs---functions,
applications, substitution---arise as shadows of these geometric
operations. In particular, {\sc spc} recovers dependent type
theory as a normal-order limit of geometric evaluation.

\subsection{Spheres as Computational Regions}

A sphere $S\subset P$ is a computational region whose internal state
is not given by a symbol assignment, but by its geometric
configuration: its embedding $i_S$, its induced metric, and its
phase structure. The computational significance of $S$ depends on
the normal bundle $N_{S/P}$ and on the entropy gradient
$\nabla_S S$, which together determine how information flows across
its boundary. Whereas conventional representations encode data in a
fixed vector space, {\sc spc} treats data as an instance of geometry.

The nesting relation between spheres induces a partial order:
$S\prec S'$ if and only if $S\subset S'$ as submanifolds of $P$.
This induces a hierarchical evaluation strategy analogous to
call-by-value or call-by-name in lambda calculus; however, the
order is not imposed syntactically, but follows from the geometry
of the plenum.

\subsection{Boundary and Coherence}

Let $\partial S$ denote the boundary of a sphere $S$. A pop must
preserve boundary coherence:
\begin{equation}
\pi|\_{\partial S} = i_S|\_{\partial S},
\end{equation}
ensuring that geometric information leaving $S$ enters the plenum
without tearing. This condition generates a coherence constraint
between nested spheres: if $S\subset S'$, then the pop on $S'$
constrains pops on $S$, and vice versa. The resulting structure is
analogous to a higher categorical boundary condition rather than a
syntactic typing rule.

\subsection{Flows and Transport}

Given a pop $\pi$, the induced flow $v_\pi$ propagates curvature and
phase into the plenum. We interpret $v_\pi$ as a computational
transport: information leaves the interior of $S$ by exerting
geometric influence on its surroundings. Unlike symbolic systems,
where output must be encoded explicitly, Spherepop transports
information by natural geometric evolution.

\section{The Pop Derivative}

The primitive computational action of {\sc spc} is not arithmetic
but collapse. A pop induces a measurable change in local structure.
Let $\mathcal{F}(S)$ denote the internal configuration of $S$. The
\emph{pop derivative} is defined by:
\begin{equation}
D_{\mathrm{pop}} \mathcal{F}(S) = 
   \lim_{\epsilon\to 0}
   \frac{\mathcal{F}(\mathrm{pop}_\epsilon(S))-\mathcal{F}(S)}{\epsilon}.
\end{equation}
Geometrically, the pop derivative measures the first-order effect of
collapse on observables of $S$. Computationally, it measures how
information is transported from the interior to the exterior.

This contrasts with symbolic differentiation, where variables are
abstract. In {\sc spc}, differentiation arises from geometric
variation. Subsequent sections relate the pop derivative to the RSVP
vector field and to thermodynamic structure.

\subsection{Boundary Measure}

If $\mu$ is a boundary measure on $\partial S$, then a pop induces
a reweighting of $\mu$:
\begin{equation}
d\mu' = \lambda_\pi \, d\mu,
\end{equation}
where $\lambda_\pi$ encodes the entropy released by collapse.
Informationally, this expresses a conservation principle: collapse
reduces internal complexity but increases boundary expressivity.

This provides a physical semantics for the \emph{activation
phenomena} of machine learning: high-dimensional patterns that
appear ``activated'' in neural networks correspond to changes in
boundary measure induced by collapse.

\section{The Merge Product}

Computation in {\sc spc} is the composite effect of multiple
collapses. Let $S_1, S_2$ denote spheres with compatible boundaries.
Define the merge product:
\begin{equation}
S_1 \diamond S_2 =
   \overline{\mathrm{pop}(S_1\cup S_2)}.
\end{equation}
In other words, we merge $S_1, S_2$ by forming their union and
collapsing the combined region into the plenum. This operator
defines the algebraic structure of {\sc spc}: composition is
collapse, and the unit is the empty sphere.

\begin{proposition}[Associativity]
If $S_1,S_2,S_3$ have mutually coherent boundaries, then
$(S_1\diamond S_2)\diamond S_3 =
 S_1\diamond(S_2\diamond S_3)$.
\end{proposition}

\begin{proof}[Sketch]
Boundary coherence ensures that the induced flows are compatible,
and collapse is determined up to homotopy by boundary. Hence the
two-step collapses are homotopic, and the resulting submanifolds
coincide up to equivalence. \qedhere
\end{proof}

Thus {\sc spc} forms a monoidal category whose objects are spheres
and whose morphisms are pops. The merge product defines horizontal
composition. As we will show, this category admits a functor to a
derived category of RSVP fields, endowing computation with physical
content.

\section{Derived Geometry}

The plenum $P$ supports scalar, vector, and entropy fields
$(\Phi,v,S)$, giving it the structure of a smooth manifold with
thermodynamic state. Spheres inherit the induced structure by
pullback. Pops generate variations in this structure, subject to
conservation laws derived from RSVP.

Let $\mathcal{M}$ be the derived moduli space of spheres in $P$.
Then {\sc spc} calculations take place in
$\mathbf{D}(\mathcal{M})$, a derived category whose objects encode
collapse histories. Each pop contributes a morphism in
$\mathbf{D}(\mathcal{M})$, and the merge product lifts to a derived
tensor product.

\subsection{Shifted Symplectic Structure}

Spherepop is naturally equipped with a shifted symplectic form
$\omega[-1]$ characteristic of derived field theories. The pop
derivative induces a Hamiltonian flow on $\mathbf{D}(\mathcal{M})$,
ensuring that computation follows a physically meaningful trajectory
in moduli space. This gives {\sc spc} a rigorous geometric semantics
analogous to the role Hamiltonian structure plays in classical mechanics.

\section{The Spherepop Calculus}
\label{sec:spc}

We now formalize {\sc spc} as a computational calculus. The
constructors introduced above admit a syntactic presentation
analogous to term construction in lambda calculus or primitive
recursive functions. Their semantics, however, is inherently
geometric.

\subsection{Syntax}

Let $\mathcal{S}$ denote the class of spheres in $P$. We write
$\lvert \mathcal{S} \rvert$ for the underlying set of all spheres
(including the empty sphere). The syntax of {\sc spc} is given by
the following grammar:
\begin{align}
  t ::= \; & S 
   \;\mid\; \mathrm{pop}(t)
   \;\mid\; t_1 \diamond t_2
   \;\mid\; t[\Phi,v,S],
\end{align}
where $S\in \mathcal{S}$ is a geometric region, $\mathrm{pop}(t)$
denotes collapse (Sec.~2), $t_1\diamond t_2$ is the merge product
(Sec.~3), and $t[\Phi,v,S]$ denotes evaluation with respect to the
ambient RSVP fields. The substitution $t[\Phi,v,S]$ marks the
extension of $t$ into the plenum geometry.

Note that $t$ is not a symbolic expression, but a formal reference
to geometric structure. Unlike symbolic calculi, where syntax
precedes interpretation, {\sc spc} syntax inherits its meaning from
geometry.

\subsection{Operational Semantics}

Evaluation is effected by geometric evolution. Let $\llbracket t
\rrbracket$ denote the semantic value of $t$; then:
\begin{align}
\llbracket S \rrbracket &= S,\\[4pt]
\llbracket \mathrm{pop}(t) \rrbracket 
 &= \mathrm{pop}(\llbracket t \rrbracket),\\[4pt]
\llbracket t_1 \diamond t_2 \rrbracket
 &= \mathrm{pop}(\llbracket t_1 \rrbracket
    \cup \llbracket t_2 \rrbracket),\\[4pt]
\llbracket t[\Phi,v,S] \rrbracket 
 &= (t,\Phi,v,S),
\end{align}
where $(t,\Phi,v,S)$ denotes the geometric configuration obtained by
gluing the RSVP fields to the sphere structure of $t$. This is not a
valuation but a geometric embedding; evaluation produces a physical
configuration, not a symbolic value.

\subsection{Typing}
\label{subsec:typing}

A typing judgment $\Gamma \vdash t : \tau$ assigns $t$ a geometric
type $\tau$. Types correspond to geometric moduli:
\begin{equation}
\tau ::= \mathbf{Sphere}
  \;\mid\; \mathbf{Pop}
  \;\mid\; \mathbf{Merge}
  \;\mid\; \mathbf{RSVP}.
\end{equation}
Typing rules follow directly from the constructors:
\begin{align}
\infer[]{\Gamma\vdash S:\mathbf{Sphere}}{} \qquad
\infer[]{\Gamma\vdash \mathrm{pop}(t):\mathbf{Pop}}
      {\Gamma\vdash t:\mathbf{Sphere}}\\[6pt]
\infer[]{\Gamma\vdash t_1\diamond t_2:\mathbf{Merge}}
      {\Gamma\vdash t_1:\mathbf{Sphere} & \Gamma\vdash
       t_2:\mathbf{Sphere}}\\[6pt]
\infer[]{\Gamma\vdash t[\Phi,v,S]:\mathbf{RSVP}}
      {\Gamma\vdash t:\mathbf{Sphere}}.
\end{align}
Typing ensures that evaluation is geometrically coherent. In
contrast with symbolic calculi, where typing is syntactic, here
types encode physical admissibility. The RSVP fields enforce
thermodynamic constraints determining whether a pop is permissible.

\subsection{Normal Form}

Evaluation terminates when no further collapse is possible. A term
$t$ is in \emph{normal form} if $\mathrm{pop}(t)=t$ up to homotopy.
Informally, all collapsible geometry has been discharged into the
plenum. In practice, normal forms represent boundary configurations
with no remaining internal structure.

\begin{proposition}[Confluence]
If $t\Rightarrow t'$ and $t\Rightarrow t''$ by two distinct
evaluation sequences, then there exists a common $u$ such that
$t'\Rightarrow u$ and $t''\Rightarrow u$.
\end{proposition}

\begin{proof}[Sketch]
Since $\diamond$ forms an associative product and collapse is
determined by boundary geometry, distinct evaluation orders differ
only by homotopy of intermediate submanifolds. Therefore computation
is confluent up to geometric equivalence, and normal forms are
unique up to homotopy. \qedhere
\end{proof}

Confluence implies that Spherepop admits a consistent notion of
semantic equivalence: two terms are equal if their geometric
realizations are homotopic. This guarantees well-definedness of
computation independent of evaluation strategy.

\subsection{Computational Adequacy}

Let $NF(t)$ denote the normal form of $t$. The calculus is
computationally adequate if $\llbracket NF(t)\rrbracket$ captures the
final geometric state of computation. Adequacy follows immediately
from confluence: every evaluation path converges to an equivalent
normal form.

\begin{theorem}[Adequacy]
For all $t$, $\llbracket t \rrbracket = \llbracket NF(t)\rrbracket$.
\end{theorem}

Thus evaluation in {\sc spc} yields a unique geometric outcome. This
constitutes the computational semantics of Spherepop calculus.

\section{Geometric Semantics}
\label{sec:geom}

The Spherepop calculus admits a semantic interpretation in terms of
derived geometry on the plenum. In contrast with symbolic calculi,
whose semantics are given by valuation mappings from expressions to
sets, {\sc spc} interprets computation as geometric evolution
constrained by RSVP fields $(\Phi,v,S)$. Every term $t$ evaluates to
a geometric configuration, and the meaning of $t$ is precisely this
configuration.

\subsection{Semantic Function}

Define the semantic function
\begin{equation}
\mathcal{J} : \mathbf{SPC} \to \mathbf{DG}(P)
\end{equation}
from the syntactic category $\mathbf{SPC}$ into the derived geometric
category $\mathbf{DG}(P)$ whose objects are geometric configurations
of spheres, boundary conditions, and RSVP fields, and whose morphisms
are pop-induced transformations. Following the evaluation rules of
Sec.~\ref{sec:spc}, we set
\begin{align}
\mathcal{J}(S) &=
  (S,i_S^\*\Phi,i_S^\*v,i_S^\*S),\\[4pt]
\mathcal{J}(\mathrm{pop}(t))
 &=
 \mathrm{pop}(\mathcal{J}(t)),\\[4pt]
\mathcal{J}(t_1\diamond t_2)
 &=
 \mathrm{pop}(\mathcal{J}(t_1)
          \cup \mathcal{J}(t_2)),\\[4pt]
\mathcal{J}(t[\Phi,v,S])
 &=
 (\mathcal{J}(t),\Phi,v,S).
\end{align}
Thus semantics is evaluation; nothing is interpreted symbolically.
The semantic value of $t$ is its geometric realization.

\subsection{Derived Category}

Let $\mathcal{M}$ denote the moduli space of spheres in $P$. Pops
generate a family of morphisms in $\mathcal{M}$; their compositions
form the hom-sets. The geometric semantics of {\sc spc} is naturally
cast in the derived category
\begin{equation}
\mathbf{D}(\mathcal{M}),
\end{equation}
whose objects are spheres with collapse history and whose morphisms
are homotopy classes of pop-induced flows. The enrichment by RSVP
fields encodes thermodynamic structure, and the induced Hamiltonian
flow provides a canonical dynamical semantics.

\subsection{Hamiltonian Semantics}

The pop derivative (Sec.~2) induces a Hamiltonian vector field on
$\mathbf{D}(\mathcal{M})$:
\begin{equation}
X_{\mathrm{pop}} = 
\iota_{D_{\mathrm{pop}}}\,\omega[-1],
\end{equation}
where $\omega[-1]$ is the shifted symplectic form associated with the
RSVP fields. Hamilton's equations govern the geometric evolution of
collapse:
\begin{equation}
\dot{t} = X_{\mathrm{pop}}(t),
\end{equation}
expressing computation as a symplectic flow. Thus computational
dynamics obey conservation constraints and thermodynamic principles.

\subsection{Thermodynamics and Entropy}

The RSVP entropy $S$ determines the dissipation of collapse. Let
$\Delta S(\mathrm{pop})$ denote the entropy released during collapse.
Boundary coherence implies that
\begin{equation}
\Delta S(\mathrm{pop}) =
\int_{\partial S}
\langle v_\pi,n\rangle\, d\mu,
\end{equation}
where $n$ is the outward normal and $\mu$ is boundary measure. Thus
computation respects thermodynamic balance: internal complexity is
reduced while boundary expressivity increases. Computation is
therefore irreducibly thermodynamic.

\subsection{Functoriality}

Since merge is associative and pop is compatible with RSVP fields,
the semantic function $\mathcal{J}$ is a monoidal functor:
\begin{equation}
\mathcal{J}(t_1\diamond t_2)
\simeq
\mathcal{J}(t_1)\otimes \mathcal{J}(t_2),
\end{equation}
where $\otimes$ is derived tensor product in $\mathbf{D}(\mathcal{M})$.
Functoriality ensures that computational composition corresponds to
geometric composition.

\subsection{Equivalence Up To Homotopy}

Confluence (Sec.~4) implies that any two evaluation sequences of $t$
produce homotopy-equivalent geometric configurations. Thus semantic
equality is homotopy equivalence:
\begin{equation}
t_1 = t_2
\qquad\text{iff}\qquad
\mathcal{J}(t_1)\simeq\mathcal{J}(t_2).
\end{equation}
This makes {\sc spc} a homotopy-complete model of computation. Its
meaning is geometric, not symbolic.

\section{Computation as Geometry}

In conventional formal systems, computation is an extrinsic process:
expressions transform other expressions, and meaning is assigned
after the fact by a semantic map. In {\sc spc}, computation is
intrinsic: the act of evaluation is geometric evolution. The
relevant quantities are curvature, entropy flow, and boundary
structure. This section develops the computational interpretation of
collapse and merge.

\subsection{Activation as Boundary Curvature Change}

Let $\kappa(\partial S)$ denote the mean curvature of the sphere
boundary induced by the embedding $i_S$. A pop modifies boundary
curvature by collapsing internal geometry into the surrounding
plenum. Define the activation operator
\begin{equation}
\mathrm{Act}(S) = \Delta \kappa(\partial S)
 = \kappa(\partial S') - \kappa(\partial S),
\end{equation}
where $S'= \mathrm{pop}(S)$. This differential measures the extent to
which collapse increases boundary expressivity. In neural networks a
similar quantity is encoded by nonlinear activation functions; here
it arises geometrically.

The activation of $t$ is thus neither a symbolic operation nor a
nonlinear transformation on embedding vectors, but a change in
geometric response at the boundary. Activation is a curvature
phenomenon.

\subsection{Propagation as Derived Transport}

Let $v_\pi$ be the vector field generated by pop $\pi$, as in
Sec.~2. Then the propagation of information is given by the solution
$\gamma(t)$ of
\begin{equation}
\dot{\gamma}(t) = v_\pi(\gamma(t)).
\end{equation}
Thus propagation is Hamiltonian transport in $\mathbf{D}(\mathcal{M})$.
Importantly, no symbolic routing is required. Information flows
because geometry evolves.

\subsection{Composition as Colimit}

Let $\mathcal{J}$ be the semantic functor of Sec.~5. The semantic
value of merge is the derived colimit in $\mathbf{D}(\mathcal{M})$:
\begin{equation}
\mathcal{J}(t_1\diamond t_2)
 \simeq \mathrm{colim}\{\mathcal{J}(t_1),\mathcal{J}(t_2)\}.
\end{equation}
Thus computational composition in {\sc spc} corresponds to forming
the colimit of geometric configurations, rather than composing
symbolic functions. All composition is geometric colimit.

\subsection{Hierarchical Control}

Nested spheres induce hierarchical control flows. A sphere $S$
embedded in $S'$ inherits boundary conditions from $S'$ and imposes
constraints on $S'$. This bidirectional coupling is not present in
symbolic control structures, which are unidirectional. Instead,
control arises from geometry: subregions influence enclosing
regions by emitting curvature; enclosing regions influence
subregions by shaping allowable pops. Computation is hierarchical
because geometry is nested.

\subsection{Learning as Curvature Minimization}

If $\Phi$ and $S$ denote RSVP scalar and entropy fields, then their
interaction defines an energy functional
\begin{equation}
\mathcal{E}(S) = \int_S \Phi\, d\mathrm{vol}
  - \int_{\partial S}\! S\, d\mu.
\end{equation}
A collapse decreases internal energy while increasing boundary
entropy, hence $\mathcal{E}$ is driven toward extremal states.
Learning corresponds to minimizing $\mathcal{E}$ over collapse
histories, subject to admissible pop operations.

\begin{proposition}[Gradient Flow]
Let $t$ be a term and let $\gamma(s)$ denote its collapse history.
Then $\gamma$ follows a gradient-like descent:
\begin{equation}
\frac{d}{ds}\mathcal{E}(\gamma(s))\le 0,
\end{equation}
with equality only at normal form.
\end{proposition}

\begin{proof}[Sketch]
Collapse reduces internal complexity and increases boundary entropy.
These changes strictly decrease $\mathcal{E}$ unless the boundary is
already maximally expressive. Thus normal forms correspond to
critical points of $\mathcal{E}$.\qedhere
\end{proof}

Learning in {\sc spc} is therefore the geometric process of
minimizing an RSVP energy functional. The result is a curvature
configuration, not a weight assignment.

\subsection{Computation as Thermodynamic Evolution}

Since pop operations are Hamiltonian flows constrained by entropy,
{\sc spc} inherits the thermodynamic interpretation of RSVP. A
computation is an irreversible evolution in the derived moduli space
driven by entropy release. Symbolic systems encode irreversibility by
arbitrary directionality (e.g., program counter); Spherepop obtains
it physically. Computation is literally the dissipation of internal
complexity into geometric boundary structure.

\section{Neural and Logical Computation as Geometric Shadows}

We now show that many standard computational formalisms arise as
shadows of Spherepop calculus when geometric structure is projected
onto fixed coordinate systems, discrete index sets, or thresholded
relations. These representations inherit computational power from
derived geometry, but sacrifice geometric semantics. In this sense,
conventional neural and logical systems may be understood as
extrinsic coordinatizations of collapse phenomena.

\subsection{Linearization and Coordinate Projection}

Let $S\subset P$ be a sphere with boundary $\partial S$ and let
$\{e_i\}$ denote a choice of local coordinate frame along $\partial
S$. Evaluation of a term $t$ at the boundary induces a coordinate
representation:
\begin{equation}
x_i = \langle n,e_i\rangle,
\end{equation}
where $n$ is the outward normal of $\partial S$. These coordinates
form a vector $x=(x_i)$, which serves as a local linear shadow of
boundary structure. We emphasize that this coordinate arises by
choice, not by intrinsic computational necessity. The internal
geometry is discarded in favor of a linear chart.

\subsection{Neural Computation as Local Pop Approximation}

A neural layer consists of an affine-linear transformation followed
by a nonlinear activation. In {\sc spc}, collapse induces curvature
change at the boundary (\S 6.1). Linearization of the pop around a
coordinate chart yields
\begin{equation}
x' = W x + b,
\end{equation}
where $W$ and $b$ capture the first-order dependence of boundary
curvature on ambient RSVP fields. The nonlinear activation is
similarly obtained by thresholding the curvature differential, e.g.,
$\mathrm{Act}(S) = \Delta\kappa(\partial S)$. Thus a neural layer is
a first-order Taylor approximation of collapse.

\begin{proposition}[Neural Shadow]
Every neural layer is the shadow of a Spherepop collapse in a local
coordinate chart, obtained by linearization of boundary curvature and
thresholding of geometric response.
\end{proposition}

Thus neural computation expresses collapse phenomena extrinsically.
We call this the \emph{neural shadow} of {\sc spc}.

\subsection{Message Passing and Graph Structure}

Graph neural networks propagate messages along edges of a graph. In
{\sc spc}, geometry induces adjacency by boundary proximity: two
points share an ``edge'' if their boundaries interact under collapse.
Define adjacency
\begin{equation}
A_{uv} = 
\mathbf{1}\{\partial S_u\cap \partial S_v\ne\emptyset\}.
\end{equation}
Then pop propagation becomes message passing:
\begin{equation}
x'_u =
\sum_v A_{uv} f(x_v),
\end{equation}
which recovers graph convolution as the projection of Hamiltonian
transport into adjacency space. Again, graph structure is a
coordinate artifact of geometric interaction.

\subsection{Logical Inference as Thresholded Collapse}

Let $S_A,S_B$ represent spheres associated with logical predicates
$A,B$. If collapses of $S_A$ and $S_B$ jointly activate curvature at
the boundary of $S_C$, then we obtain a logical implication:
\begin{equation}
(A\wedge B)\Rightarrow C.
\end{equation}
Thresholding curvature defines Boolean truth:
\begin{equation}
C = 
\mathbf{1}\{\mathrm{Act}(S_C) > \tau\}.
\end{equation}
Thus logical inference is the discretization of collapse-mediated
interaction. Its syntactic form obscures the geometric mechanism.

\subsection{Probabilistic Inference as Boundary Measure}

Let $\mu(\partial S)$ be the boundary measure induced by RSVP
entropy. Collapse updates $\mu$, generating posterior distributions:
\begin{equation}
\mu'(\partial S) =
\frac{1}{Z}\exp(-\mathcal{E}(S)).
\end{equation}
Thus Bayesian update is recovered as a Gibbs-type reweighting of
boundary measure. Probabilistic inference is a thermodynamic shadow
of geometric evolution.

\subsection{Combinatorial Computation as Degenerate Geometry}

When all spheres are degenerate and collapse is trivial, {\sc spc}
reduces to a combinatorial calculus on a discrete set of symbols.
This recovers conventional symbolic computation:
\begin{equation}
\text{collapse trivial}
\quad\Longrightarrow\quad
\mathbf{SPC} \simeq \mathbf{Symb}.
\end{equation}
Hence symbolic computation arises as the extreme case in which
geometric semantics vanish.

\subsection{Summary}

Neural, logical, and probabilistic formalisms appear as coordinate
projections of Spherepop into particular representational spaces:
local vector spaces, adjacency graphs, or Boolean domains. Collapse
is the intrinsic phenomenon; layers, message passing, and logical
implication are extrinsic shadows. Computation is geometric; neural
and logical systems describe its projections.

\section{Equivalence to Lambda Calculus and Functional Paradigms}

Classical theories of computation rely on symbolic formalisms, the
most canonical being lambda calculus. We now show that {\sc spc}
contains lambda abstraction and application as special cases of
geometric embedding and collapse. Consequently, functional
programming paradigms appear as syntactic shadows of geometric
evaluation.

\subsection{Lambda Abstraction as Geometric Scope}

Let $\lambda x.\;t$ denote a lambda term with bound variable $x$.
In {\sc spc}, abstraction corresponds to introducing a sub-sphere
whose boundary isolates the scope of the variable. Let $S_x$
denote a sphere encoding the scope of $x$ and let $S_t$ encode
the geometry of $t$. Then lambda abstraction is represented by
the nested sphere
\begin{equation}
\Lambda_x(t) := S_x \prec S_t,
\end{equation}
where nesting expresses dependency of $t$ on $x$. The boundary of
$S_x$ plays the role of the abstraction boundary.

\subsection{Application as Collapse and Merge}

Application $(\lambda x.\,t)\;u$ substitutes $u$ for $x$ in $t$. In
{\sc spc}, substitution is executed by collapsing the $S_u$ sphere
into the abstraction sphere $S_x$ and then merging with $S_t$:
\begin{equation}
\mathrm{App}(t,u) :=
\mathrm{pop}(S_x\diamond S_u)\diamond S_t.
\end{equation}
Operationally, application is the geometric process of replacing the
scope of $x$ with the geometry of $u$. Collapse performs
substitution; merge performs evaluation.

\begin{proposition}[Beta Reduction]
Let $t,u$ be terms, and assume $x$ is bound in $t$. Then evaluation
of $(\lambda x.\,t)\;u$ by {\sc spc} yields a term homotopic to
$t[u/x]$, the standard beta-reduced form.
\end{proposition}

\begin{proof}[Sketch]
Collapse of $S_u$ into $S_x$ substitutes internal geometry; merge
into $S_t$ propagates effects into the body of $t$. Confluence
ensures uniqueness up to homotopy.\qedhere
\end{proof}

Thus beta-reduction corresponds to geometric collapse, and eta- and
alpha-equivalences correspond to homotopies of sphere boundaries.

\subsection{Higher-Order Functions}

Since spheres may contain sub-spheres, and collapses may operate on
regions containing collapses, {\sc spc} directly supports
higher-order computation. A higher-order function $F:(A\to B)\to C$
is represented by a sphere $S_F$ whose boundary admits pop operations
on sub-spheres corresponding to functions of type $A\to B$.
Evaluation of $F$ on such a function amounts to collapse within
$S_F$.

\subsection{Functional Paradigms as Geometric Evaluation}

Let $\mathbf{LC}$ denote the category of lambda terms up to beta
equivalence. The mapping
\begin{equation}
\mathcal{L}:\mathbf{LC}\to\mathbf{SPC}
\end{equation}
embedding lambda calculus into Spherepop identifies abstraction with
nested spheres and application with collapse and merge. Thus
functional programming paradigms---including higher-order functions,
closures, and currying---arise as syntactic presentations of geometric
evaluation.

\subsection{Universality}

Lambda calculus is known to be Turing complete. Since {\sc spc}
contains lambda calculus via the embedding $\mathcal{L}$, it is
computationally universal:
\begin{equation}
\mathbf{LC}\hookrightarrow\mathbf{SPC}
\quad\Longrightarrow\quad
\mathbf{SPC}\text{ is Turing complete}.
\end{equation}
Conversely, $\mathbf{SPC}$ extends lambda calculus by giving it
geometric semantics, higher-dimensional structure, and thermodynamic
interpretation.

\subsection{Geometric Generalization}

As symbolic calculi capture only the degenerate case in which
geometric semantics are suppressed, lambda calculus appears as the
zero-dimensional shadow of {\sc spc}. In this view, functional
programming corresponds to collapse restricted to purely syntactic
constructors. The geometric model is strictly stronger: it subsumes
functional paradigms and provides additional structure unavailable in
symbolic formalisms.

\subsection{Summary}

Spherepop calculus strictly extends lambda calculus: it contains
abstraction and application as geometric primitives, recovers
beta-reduction as collapse, and supports higher-order functions
intrinsically. Functional programming is thus the syntactic shadow of
geometric computation.

\section{Tensor Logic as Flat Boundary Projection}
\label{sec:tl}

Having established that {\sc spc} subsumes classical computation
(\S8) and that neural and logical systems appear as geometric
shadows (\S7), we now show that Tensor Logic arises when Spherepop
is restricted to flat boundary geometry and evaluated through local
linear projections. In this regime, collapse becomes tensor
contraction, and merge becomes Einstein summation.

\subsection{Preliminaries}

Tensor Logic (TL) models reasoning in terms of tensor contractions
among embedding vectors. Its central operations are projection,
einsum, and threshold. We interpret these as flattened forms of
collapse and merge over boundary charts on $\partial S$. The
semantics of {\sc spc} therefore induces Tensor Logic by geometric
degeneration: TL is the projection of Spherepop onto linear
boundary coordinates with vanishing curvature and suppressed
interior.

\subsection{Flat Boundary Condition}

Let $S\subset P$ be a sphere with boundary $\partial S$ and curvature
$\kappa(\partial S)$ (Sec.~6). If $\kappa(\partial S)=0$ for all $S$
participating in a computation, then boundary geometry is flat, and
collapse simplifies to selection and contraction of coordinates.
Formally,
\begin{equation}
\kappa(\partial S)=0
\quad\Longrightarrow\quad
\mathrm{pop}(S)
 = \mathrm{Proj}(\partial S),
\end{equation}
where $\mathrm{Proj}$ denotes projection onto a fixed boundary chart.
Thus Tensor Logic corresponds to zero-curvature evaluation.

\subsection{Merge as Einstein Summation}

Under flat boundary conditions, merge induces contraction:
\begin{equation}
\mathcal{J}(t_1\diamond t_2)
  \simeq \sum_{i,j} 
        T^{(1)}_{i\ell}
        T^{(2)}_{\ell j},
\end{equation}
recovering the familiar Einstein summation rule. The summation index
$\ell$ corresponds to the internal region collapsed by pop. In
general geometry, contraction is replaced by geometric merge; the
tensor operation is merely the flat projection.

\subsection{Linearization of Collapse}

Let $\mathrm{Act}(S)=\Delta\kappa(\partial S)$ be the activation of
collapse (Sec.~6.1). Under flat boundary conditions and local
linearization,
\begin{equation}
\mathrm{Act}(S)\approx
   W x,
\end{equation}
with $x$ the boundary coordinate of the incoming sphere and $W$ the
Jacobian of pop. Thresholding yields Boolean or piecewise-linear
activation, reproducing the Tensor Logic interpretation of nonlinear
reasoning in embedding space.

\subsection{Semantic Equivalence}

Let $\mathbf{TL}$ denote the category of Tensor Logic expressions up
to the usual equivalence relations. Then Spherepop induces a
functor
\begin{equation}
\mathcal{T}:\mathbf{SPC}\to\mathbf{TL}
\end{equation}
by restricting to flat boundaries and linear charts. Moreover,
collapse and merge reduce to contraction and einsum, giving an
equivalence
\begin{equation}
\mathbf{SPC}\big|\_{\kappa=0}
  \simeq 
\mathbf{TL}.
\end{equation}
Thus Tensor Logic is not a foundational theory; it is the
zero-curvature projection of Spherepop.

\subsection{Domingos as Local Linear Shadow}

In Domingos’ formulation, reasoning is expressed through linear
operations in embedding space with logical structure encoded by
tensor transformations. This corresponds precisely to the local
linearization of collapse at flat boundaries. As such, the Tensor
Logic account of reasoning appears as the geometrically flattened
restriction of {\sc spc}: reasoning in embedding space is the
twistor projection of collapse.

\subsection{Algebraic Equivalence Under Boundary Flattening}

Let $\mathcal{J}$ be the semantic functor of Sec.~\ref{sec:geom}.
We restrict $\mathcal{J}$ to flat boundaries by imposing the
constraint $\kappa(\partial S)=0$ on all spheres $S$. Algebraically,
collapse becomes linear projection, and merge becomes bipartite tensor
contraction. Let
\begin{equation}
\mathcal{J}_0 := \mathcal{J}\big|_{\kappa=0}.
\end{equation}
Then $\mathcal{J}_0$ induces a functor
$\mathcal{J}_0 : \mathbf{SPC} \to \mathbf{TL}$ such that
\begin{equation}
\mathcal{J}_0(t_1\diamond t_2)
   = \mathcal{J}_0(t_1)\,\oplus\,\mathcal{J}_0(t_2),
\end{equation}
where $\oplus$ denotes the tensor connective of Tensor Logic.

\begin{proposition}[Algebraic Equivalence]
Under flat boundary conditions, collapse and merge induce the algebra
of tensor connectives in Tensor Logic.
\end{proposition}

\subsection{Categorical Correspondence via Monoidal Restriction}

Let $\mathbf{SPC}$ be the monoidal category generated by spheres and
collapse with monoidal product $\diamond$. Let $\mathbf{TL}$ be the
monoidal category whose objects are embedding tensors and whose
arrows are tensor contractions. The flattening functor
$\mathcal{T}$ of Sec.~9 induces a monoidal functor
\begin{equation}
\mathcal{T}:\mathbf{SPC}\to\mathbf{TL}
\end{equation}
with
\begin{equation}
\mathcal{T}(t_1\diamond t_2)
 = \mathcal{T}(t_1)\otimes \mathcal{T}(t_2).
\end{equation}
Hence Tensor Logic is the monoidal reflection of Spherepop under
curvature suppression.

\begin{theorem}[Monoidal Equivalence]
If all collapses are flat and boundary curvature vanishes, then
$\mathcal{T}$ induces a monoidal equivalence
\[
\mathbf{SPC}\big|\_{\kappa=0}
  \simeq
\mathbf{TL}.
\]
\end{theorem}

\subsection{Computational Equivalence Through Local Linearization}

Tensor Logic represents computation by manipulating embeddings through
tensor contractions and threshold operations. Under local
linearization of collapse (Sec.~6),
\begin{equation}
\mathrm{pop}(S) \approx W x + b,
\end{equation}
and merge induces tensor multiplication. Let
$\mathrm{TL}(t)$ denote the Tensor Logic interpretation of $t$. Then
local linearization yields
\begin{equation}
\mathrm{TL}(t)
 = \mathcal{J}_0(t)[W,b].
\end{equation}
Thus every Tensor Logic computation is the local linearization of
a Spherepop evaluation.

\begin{corollary}[Computational Equivalence]
Every Tensor Logic computation is executable as the first-order
linear approximation of a Spherepop collapse.
\end{corollary}

\subsection{Model-Theoretic Interpretation}

Tensor Logic inherits a model-theoretic semantics based on embeddings
and projections. Spherepop inherits a geometric semantics based on
collapse and boundary curvature. Under flat boundary conditions,
the semantic equations coincide:
\begin{equation}
\mathcal{J}_0(t) = \mathrm{TL}(t).
\end{equation}
Thus Tensor Logic may be regarded as the model theory of
zero-curvature Spherepop. Model-theoretic entailment corresponds to
the evaluation of derived collapse under projection.

\begin{proposition}
Model-theoretic validity in Tensor Logic coincides with semantic
validity of {\sc spc} under flat boundary evaluation.
\end{proposition}

\subsection{Operational Collapse vs. Einsum}

The operational semantics of {\sc spc} interprets computation as the
geometric act of collapse. Tensor Logic interprets computation as
einsum over embeddings. Under curvature suppression,
\begin{equation}
\mathrm{pop}(t_1\diamond t_2)
  = \mathrm{einsum}(t_1,t_2),
\end{equation}
making einsum the degenerate operational rule of collapse.
Consequently, Tensor Logic is not a novel algebraic foundation, but a
special operational regime of Spherepop.

\subsection{Summary}

Tensor Logic emerges from Spherepop when curvature is suppressed,
interior geometry is collapsed to zero thickness, and merge is
linearized to tensor contraction. Einsum and projection are special
cases of geometric merge and collapse. Reasoning in Tensor Logic
represents the flat boundary shadow of geometric computation.

\section{Amplitwistors and Geometric Computation}
\label{sec:ampli}

Tensor representations flatten geometric structure into local linear
coordinates. In contrast, twistor geometry provides a holomorphic
representation in which null directions of spacetime are encoded as
points of a complex projective space. Amplitwistors extend this idea
to scattering amplitudes, providing a nonlocal representation of
interactions. In this section we show that Spherepop calculus admits
an amplitwistor interpretation in which collapse becomes a nonlocal
update over twistor fields.

\subsection{Twistor Background}

Let $(M,g)$ be a four-dimensional Lorentzian manifold. The (projective)
twistor space $PT$ is the complex manifold parameterizing null
geodesics in $M$. Each point of $PT$ corresponds to a null direction
in $M$, and each null geodesic corresponds to a complex projective
line in $PT$. This construction transforms local geometric structure
on $M$ into holomorphic data on $PT$, enabling computation on a
nonlocal complex manifold.

\subsection{Amplitwistors}

Amplitwistors extend the twistor description to scattering theory,
encoding momenta and polarization data in complex geometry. In the
amplitwistor formalism, scattering amplitudes are computed by
integrating holomorphic differential forms over complex submanifolds
of an amplitwistor space. The resulting formalism provides a
nonlocal, coordinate-free representation of multi-particle
interaction.

\subsection{Spherepop in Twistor Coordinates}

Let $S\subset P$ be a sphere encoding local geometry. Under
twistorization, the geometry of $S$ is encoded as a complex locus
$Z_S\subset PT$. Collapse corresponds to modifying the complex
structure on the locus, and merge corresponds to composing such
modifications:
\begin{equation}
\mathcal{J}(t)
 \mapsto
F(Z_S).
\end{equation}
Thus Spherepop computations can be performed directly in twistor
space by manipulating complex loci.

\subsection{Nonlocality}

The twistor representation makes explicit the nonlocal structure
implicit in Spherepop collapse: the effect of collapse on $P$ may
propagate along null directions in $PT$. This provides a natural
interpretation of nonlocal computational effects, which do not
require explicit message passing or symbolic communication. In
twistor space, computation is inherently nonlocal.

\subsection{Holomorphic Collapse}

Pop operations act by modifying complex structure on $PT$. Let
$D_{\mathrm{pop}}$ be the collapse derivative of Sec.~2. Then
\begin{equation}
D_{\mathrm{pop}}
  \mapsto
\bar{\partial}\text{-operator on }PT,
\end{equation}
identifying collapse with a holomorphic differential operator. Thus
Spherepop inherits a holomorphic semantics in amplitwistor space.

\subsection{Comparison with Tensor Logic}

Tensor Logic linearizes geometric computation in local coordinates,
disregarding nonlocality and complex structure. Amplitwistor
coordinates, in contrast, preserve the global geometry of collapse
and provide a natural arena for nonlocal computation. Rather than
reducing computation to tensor contraction, amplitwistors lift it
to holomorphic geometry. Tensor Logic is a flattening; amplitwistors
are a lifting.

\subsection{Summary}

Amplitwistors provide a complex-analytic representation in which
Spherepop collapse is expressed as a holomorphic differential
operator. Computation becomes nonlocal by construction, and geometric
structure is preserved. In contrast to Tensor Logic, amplitwistors
respect the manifold geometry underlying {\sc spc}.

\section{Collapse as Twistor Scattering}
\label{sec:twistorsc}

Spherepop collapse acts locally on $P$ but propagates globally in $PT$.
This resemblance to scattering---local interaction yielding global
correlations---is made precise in amplitwistor geometry. In this
section we interpret pop as a twistor scattering operation modifying
holomorphic data along complex contours.

\subsection{Scattering Interpretation}

Consider a collapse $\mathrm{pop}(S)$ in $P$. Under the twistor
mapping of Sec.~\ref{sec:ampli}, $S$ maps to a locus $Z_S\subset PT$.
Collapse modifies near-boundary geometry, hence modifies $Z_S$.
Let $\Delta Z_S$ denote the resulting change in complex structure.
The geometric change induced by collapse resembles the insertion of
a vertex operator or interaction term in scattering theory.

\begin{equation}
\mathrm{pop}(S)
 \longmapsto
Z_S\mapsto
Z_S + \Delta Z_S.
\end{equation}

Thus collapse generates a twistor scattering event.

\subsection{Holomorphic Scattering Operator}

Let $\mathcal{A}$ denote the amplitwistor scattering operator acting
on complex loci. Then pop induces an operator
\begin{equation}
\mathcal{A}_{\mathrm{pop}}:Z_S\to Z_S',
\end{equation}
which modifies complex structure by holomorphic deformation. In
local coordinates this corresponds to insertion of an operator
$\mathcal{O}_{\mathrm{pop}}$ acting on a holomorphic differential
form on $PT$.

\subsection{Contour Deformation}

Scattering amplitudes are computed by integrating holomorphic forms
over contours in amplitwistor space. Collapse induces deformation of
these contours, thus modifying amplitudes:
\begin{equation}
\int_\Gamma \Omega
 \quad\mapsto\quad
\int_{\Gamma'} \Omega.
\end{equation}
Hence collapse corresponds to contour modification in twistor space,
in analogy with the deformation of scattering contours in QFT.

\subsection{Merge as Composite Scattering}

Merge $\diamond$ combines geometric regions in $P$; in amplitwistor
geometry this becomes gluing of the corresponding complex loci:
\begin{equation}
Z_{t_1\diamond t_2}
   = \mathrm{Glue}(Z_{t_1},Z_{t_2}).
\end{equation}
If collapse corresponds to scattering, merge corresponds to composition
of scattering processes. A composite pop sequence represents a
multi-vertex twistor interaction.

\subsection{Holomorphic Action Functional}

Let $\Omega$ be the holomorphic volume form on amplitwistor space.
Define the action
\begin{equation}
\mathcal{S}[Z_S]
 = \int_{Z_S}\Omega.
\end{equation}
Collapse transforms $Z_S$ and thus changes $\mathcal{S}$. The result
is a twistor scattering amplitude for collapse. This gives
Spherepop an amplitwistor action principle.

\subsection{Summary}

Collapse in $P$ induces scattering in $PT$: pop acts as a holomorphic
operator, merge corresponds to composite interactions, and geometric
evaluation becomes contour deformation. Pop is scattering.

\section{Amplitwistor Integration and Global Interactions}
\label{sec:ampint}

Having interpreted collapse as a scattering process in amplitwistor
space (Sec.~\ref{sec:twistorsc}), we now integrate these local
scattering events to obtain global computational behavior. In
traditional scattering theory, amplitudes are obtained by integrating
local interaction data over contours in a complex manifold. The
analogous construction in {\sc spc} integrates collapse histories
over amplitwistor space to produce global geometric inference.

\subsection{Collapse Histories}

Let $\gamma$ be a collapse history for a term $t$. Under the
amplitwistor representation of Sec.~\ref{sec:ampli}, $\gamma$ maps to
a family of complex loci $\{Z_{\gamma(s)}\}_{s\in[0,1]}$. Collapse
generates a trajectory in amplitwistor space, and global computation
corresponds to integrating holomorphic data along this trajectory.

\subsection{Global Amplitwistor Integral}

Let $\Omega$ be the holomorphic measure in amplitwistor space and let
$Z_{\gamma}$ denote the twistor locus associated with collapse history
$\gamma$. Define the global computation
\begin{equation}
\mathcal{I}(t)
 = \int_{Z_\gamma}\Omega,
\end{equation}
where the integral runs over all collapse-induced loci of $t$.
Hence evaluation of a term in {\sc spc} corresponds to a holomorphic
integral in amplitwistor space.

\subsection{Nonlocality}

Unlike conventional symbolic evaluation, the integral
$\mathcal{I}(t)$ depends on the global geometry of $Z_{\gamma}$. This
encodes nonlocal correlations, as collapse of a local region in $P$
may propagate along null directions in $PT$. In this sense,
computation in the amplitwistor representation is inherently
nonlocal, independent of any explicit message passing mechanism.

\subsection{Composite Interaction}

Merge composes two regions in $P$ into a single computational unit.
Under amplitwistor integration this becomes contour gluing:
\begin{equation}
\int_{Z_{\gamma_1\diamond\gamma_2}}\Omega
 =
\int_{Z_{\gamma_1}}\Omega
  \circ
\int_{Z_{\gamma_2}}\Omega.
\end{equation}
Thus the algebra of merge operations corresponds to the algebra of
composite contour integrals, making merge a global interaction rule.

\subsection{Scattering Amplitudes as Computation}

The integral $\mathcal{I}(t)$ resembles a scattering amplitude
obtained by integrating interaction data over complex contours in
amplitwistor space. In {\sc spc}, collapse produces such interaction
data; global computation is therefore equivalent to evaluating a
scattering amplitude. Pop computes by scattering.

\subsection{Holomorphic Inference}

The global amplitwistor integral provides a principle of inference:
\begin{equation}
t_1 = t_2
\quad\Longleftrightarrow\quad
\mathcal{I}(t_1)=\mathcal{I}(t_2).
\end{equation}
Two terms are equivalent if their amplitwistor integrals coincide. In
contrast to Tensor Logic, which relies on algebraic equalities of
linear operators, {\sc spc} inherits a holomorphic criterion for
computational equivalence.

\subsection{Summary}

Global computation in Spherepop is the amplitwistor integral of
collapse histories. Merge becomes global gluing, collapse becomes
scattering, and equivalence is determined by equality of holomorphic
integrals. Inference is holomorphic, not linear-algebraic.

\section{Twistor--RSVP Coupling}
\label{sec:twistrsvp}

The amplitwistor interpretation of Spherepop provides a complex
manifold representation of collapse. The RSVP framework introduces
a scalar field $\Phi$, a vector field $v$, and an entropy field $S$
on the plenum $P$. We now couple these representations by lifting
RSVP fields to twistor space and interpreting collapse as a coupled
holomorphic--thermodynamic evolution. This yields a twistorized form
of the plenum in which computation is expressed through amplitwistor
geometry.

\subsection{RSVP Fields}

Let $P$ be the plenum manifold and $(\Phi,v,S)$ the RSVP fields. The
scalar field $\Phi$ encodes potential, the vector field $v$ encodes
flow, and the entropy field $S$ encodes thermodynamic structure.
These fields evolve according to plenum dynamics, which are not
assumed to arise from a metric variation or from Einstein equations.
The plenum $P$ carries collapse-induced dynamics determined by
thermodynamic constraints.

\subsection{Twistor Lift}

Let $\pi:PT\to P$ denote the twistor fibration. The RSVP fields admit
a twistor lift:
\begin{equation}
\Phi^{\sharp}
 = \pi^\*\Phi,\qquad
v^{\sharp}
 = \pi^\*v,\qquad
S^{\sharp}
 = \pi^\*S.
\end{equation}
Thus the RSVP fields define holomorphic data on twistor space, making
collapse a holomorphic--thermodynamic evolution.

\subsection{Coupled Collapse}

Collapse modifies boundary geometry $S\subset P$, hence modifies the
lifted RSVP fields:
\begin{equation}
(\Phi^{\sharp},v^{\sharp},S^{\sharp})
 \mapsto
(\Phi^{\sharp}+\Delta\Phi^{\sharp},\,
 v^{\sharp}+\Delta v^{\sharp},\,
 S^{\sharp}+\Delta S^{\sharp}).
\end{equation}
These modifications in $PT$ correspond to transformed thermodynamic
conditions in $P$. Collapse therefore couples amplitwistor geometry to
plenum thermodynamics.

\subsection{Twistor--Lamphrodynamic Semantics}

Let $\mathcal{A}$ be the amplitwistor scattering operator of
Sec.~\ref{sec:twistorsc}. Define the twistor--lamphrodynamic operator
\begin{equation}
\mathcal{L} :=
\mathcal{A}
  \circ
(\mathrm{lift\;RSVP}).
\end{equation}
Then the semantic value of $t$ in twistor space is
\begin{equation}
\mathcal{J}(t)
 = \mathcal{L}(Z_t),
\end{equation}
where $Z_t$ is the amplitwistor locus for term $t$. Thus {\sc spc}
evaluation becomes a lamphrodynamically constrained scattering
process on twistor space.

\subsection{Thermodynamic Curvature}

The RSVP entropy $S$ modulates curvature in $P$. Under the twistor
lift, $S^{\sharp}$ modulates complex geometry in $PT$, influencing
the holomorphic structure of collapse. This yields a curvature
representation coupling thermodynamic and complex-analytic data. The
result is a twistor-thermodynamic geometry that governs computation.

\subsection{Comparison with Tensor Logic}

Tensor Logic encodes computational transformations by linear
operators on embedding vectors. Twistor--RSVP coupling encodes
computational transformations by thermodynamically constrained
holomorphic operators. Tensor Logic discards both thermodynamics and
geometry, replacing collapse by linear algebra. Twistor--RSVP
restores the full geometric--thermodynamic semantics of computation.

\subsection{Summary}

Collapse in Spherepop induces thermodynamic evolution of RSVP fields.
Its twistor representation induces holomorphic evolution of lifted
RSVP fields in amplitwistor space. Computation therefore acquires a
twistor--thermodynamic semantics in which collapse is a coupled
holomorphic--lamphrodynamic process. Tensor Logic corresponds to the
degenerate, zero-curvature projection of this geometry.

\section{Hierarchical Twistor--RSVP Computation}
\label{sec:hier}

The coupling developed in Sec.~\ref{sec:twistrsvp} defines computation
as a thermodynamically constrained holomorphic scattering process in
amplitwistor space. In this section we introduce a hierarchical
structure based on nested spheres, which induces hierarchical
twistor--RSVP evolution. Computation propagates across scales of the
plenum and across layers of amplitwistor geometry.

\subsection{Nested Spheres and Multiscale Collapse}

Let $S\subset S'\subset P$ be nested spheres representing multiscale
regions of the plenum. Under twistorization, these map to nested loci
$Z_S\subset Z_{S'}\subset PT$. Collapse of $S$ modifies $Z_S$ and
subsequently modifies $Z_{S'}$ due to their containment relation. Thus
collapse induces multiscale twistor interactions.

\subsection{Hierarchical Lamphrodynamic Evolution}

The RSVP fields $(\Phi,v,S)$ govern local evolution in $P$.
Hierarchically nested spheres induce hierarchical lamphrodynamic
constraints:
\begin{equation}
(\Phi,v,S)_{S'} 
 \mapsto
(\Phi,v,S)_{S}
 \mapsto
(\Phi,v,S)_{\mathrm{int}},
\end{equation}
where the subscript indicates restriction to nested regions. Collapse
propagates thermodynamic constraints across scales, inducing
hierarchical computational behavior.

\subsection{Twistor Hierarchy}

The twistor lift $\pi^\*$ applies to each level of the hierarchy,
producing a hierarchy in $PT$:
\begin{equation}
Z_{\mathrm{global}}
 \supset
Z_{\mathrm{regional}}
 \supset
Z_{\mathrm{local}}.
\end{equation}
Collapse at one level modifies the lifted entropy and complex
structure at finer or coarser levels. Thus computation is both local
and global in twistor space.

\subsection{Hierarchical Scattering}

Let $\mathcal{A}$ denote the twistor scattering operator. Nested
collapse yields
\begin{equation}
\mathcal{A}_{S}
 \circ
\mathcal{A}_{S'}
 \circ
\mathcal{A}_{\mathrm{global}},
\end{equation}
representing a chain of twistor interactions. Computation is thus a
hierarchical scattering process governed by RSVP thermodynamics.

\subsection{Multiscale Integration}

The global amplitwistor integral of Sec.~\ref{sec:ampint} decomposes
into nested contributions:
\begin{equation}
\mathcal{I}(t)
 =
\int_{Z_{\mathrm{global}}}\Omega
 =
\int_{Z_{\mathrm{regional}}}\Omega
 =
\int_{Z_{\mathrm{local}}}\Omega,
\end{equation}
yielding a multiscale holomorphic inference principle. Local collapse
participates in global inference by way of the twistor hierarchy.

\subsection{Summary}

Nested spheres induce hierarchical lamphrodynamic constraints, which
lift to hierarchical twistor geometry. Collapse propagates across
scales, producing multiscale scattering events and global
holomorphic inference. Computation is therefore hierarchical in both
the plenum and twistor space.

\section{Holomorphic Entropy and Nonlocal Inference}
\label{sec:holomentropy}

Entropy in RSVP induces lamphrodynamic relaxation in the plenum. Under
the twistor lift, entropy becomes a holomorphic field in amplitwistor
space. We now show that collapse induces holomorphic entropy flow,
producing nonlocal inference by complex propagation rather than local
message passing or symbolic deduction.
\subsection{Holomorphic Entropy Field}

Let $S^{\sharp}=\pi^\*S$ be the lifted entropy field. Collapse alters
$S^{\sharp}$ along complex loci $Z_S$, inducing a holomorphic entropy
flow
\begin{equation}
\bar{\partial}S^{\sharp}
  = \Delta S^{\sharp}\, .
\end{equation}
This identifies collapse with entropy-driven evolution in twistor
space.
\subsection{Nonlocality}

The modification of $S^{\sharp}$ along $Z_S$ propagates in $PT$ along
null complex directions, making inference nonlocal by construction.
This differs fundamentally from Tensor Logic, which relies on local
embedding coordinates and linear contraction.

\subsection{Holomorphic Inference}

Inference between regions $S_A$ and $S_B$ occurs when collapse of
$S_A$ modifies $S^{\sharp}$ at $S_B$. The resulting inference relation
is given by
\begin{equation}
S_A \leadsto S_B
\quad\Longleftrightarrow\quad
\Delta S^{\sharp}(Z_{S_A})
  \longrightarrow
\Delta S^{\sharp}(Z_{S_B}).
\end{equation}
Thus inference arises through holomorphic entropy flow, not symbolic
deduction.
\subsection{Complex-Analytic Learning}

Learning in {\sc spc} (Sec.~6.5) minimizes plenum energy. Under the
twistor lift, learning corresponds to modifying $S^{\sharp}$ to
reduce holomorphic variation. Collapse (learning) therefore drives
the system toward holomorphic equilibrium states in amplitwistor
space.
\subsection{Summary}

Entropy couples collapse to a holomorphic differential operator,
producing nonlocal inference through complex propagation. Learning
minimizes entropy variation, driving the system toward holomorphic
equilibrium. Inference is holomorphic and nonlocal.
\section{Toward a Unifying Computational Physics}
\label{sec:unified}

We conclude that Spherepop calculus, when interpreted in amplitwistor
coordinates and coupled to RSVP fields, yields a computational physics
based on thermodynamically constrained holomorphic scattering.

Tensor Logic and lambda calculus arise as degenerate shadows of this
geometry, and neural and logical systems appear as coordinate
projections. Computation is geometric, thermodynamic, and holomorphic
by construction.
\subsection{Unification}

Spherepop unifies:
  \begin{enumerate}
    \item geometric collapse and merge,
    \item thermodynamic entropy flow,
    \item holomorphic scattering,
    \item hierarchical inference,
    \item functional abstraction,
    \item tensor contraction as degeneration.
  \end{enumerate}
Computation arises from geometric first principles, not symbolic
construction.
\subsection{Future Directions}

Future work includes:
  \begin{enumerate}
    \item rigorous construction of amplitwistor--RSVP scattering theory,
    \item categorical semantics of holomorphic collapse,
    \item computational completeness of twistor--RSVP evolution,
    \item coupling with physical twistor theories,
    \item experimental consequences for geometric machine learning.
  \end{enumerate}
\subsection{Conclusion}

Spherepop calculus defines computation as geometric collapse and
thermodynamic evolution, and amplitwistors provide a complex-analytic
representation of this computation. Tensor Logic emerges only as the
flat, linearized projection of this geometry. Computational physics
is therefore geometric and amplitwistor-based rather than
syntactically tensor-based.

\appendix

\section{Formal Specification of Spherepop Calculus}

\subsection{Syntax}

Let $\mathcal{S}$ be the class of admissible spheres in the plenum $P$.
Spherepop terms are expressions of the form
\[
t ::= S \;\mid\; \mathrm{pop}(t)\;\mid\; t_1\diamond t_2\;\mid\; t[\Phi,v,S],
\]
where $S\in\mathcal{S}$, $\mathrm{pop}(t)$ denotes collapse,
$t_1\diamond t_2$ denotes merge, and $t[\Phi,v,S]$ denotes evaluation
relative to the ambient fields $(\Phi,v,S)$.

\subsection{Types}

Spherepop types are given by
\[
\tau ::= \mathbf{Sphere}\;\mid\;\mathbf{Pop}\;\mid\;\mathbf{Merge}
         \;\mid\;\mathbf{RSVP}.
\]

Typing judgments $\Gamma\vdash t:\tau$ satisfy
\[
\frac{}{\,\Gamma\vdash S:\mathbf{Sphere}\,}, \qquad
\frac{\Gamma\vdash t:\mathbf{Sphere}}
     {\,\Gamma\vdash\mathrm{pop}(t):\mathbf{Pop}\,},
\]
\[
\frac{\Gamma\vdash t_1:\mathbf{Sphere}\qquad\Gamma\vdash t_2:\mathbf{Sphere}}
     {\,\Gamma\vdash t_1\diamond t_2:\mathbf{Merge}\,},
\qquad
\frac{\Gamma\vdash t:\mathbf{Sphere}}
     {\,\Gamma\vdash t[\Phi,v,S]:\mathbf{RSVP}\,}.
\]

\subsection{Operational Semantics}

Let $\llbracket t\rrbracket$ denote the geometric configuration
represented by $t$. Evaluation is defined by
\[
\llbracket S\rrbracket = S,\qquad
\llbracket \mathrm{pop}(t)\rrbracket =
\mathrm{pop}(\llbracket t\rrbracket),
\]
\[
\llbracket t_1\diamond t_2\rrbracket
  =\mathrm{pop}(\llbracket t_1\rrbracket\cup\llbracket t_2\rrbracket),
\qquad
\llbracket t[\Phi,v,S]\rrbracket = (t,\Phi,v,S).
\]
Evaluation yields geometric configurations rather than symbolic
values.

\subsection{Normal Forms}

A term $t$ is in normal form if $\mathrm{pop}(t)=t$ up to homotopy.
Confluence and adequacy (proved in the main text) imply that normal
forms are unique up to geometric equivalence.

\section{RSVP Fields and Variational Formulation}

Let $P$ be a differentiable manifold (the plenum). An RSVP structure is
specified by fields
\[
(\Phi,\,v,\,S),
\]
where $\Phi$ is a scalar potential, $v$ a vector field (interpreted as
flow), and $S$ an entropy field. Collapse and merge operations interact
with these fields through a variational principle. A standard
Lagrangian density is given by
\[
\mathcal{L}[\Phi,v,S]
 = \tfrac{1}{2}\,|\nabla\Phi|^2
   +\langle v,\nabla\Phi\rangle
   -S,
\]
where $|\nabla\Phi|^2$ denotes the metric contraction of the gradient,
and $\langle v,\nabla\Phi\rangle$ denotes flow-mediated interaction.
The entropy term $-S$ represents internal-to-boundary dissipation.

Euler--Lagrange variation yields a coupled PDE system of the form
\[
\Delta\Phi - \mathrm{div}(\Phi v)= \partial S/\partial t,
\qquad
\partial_t\,v + (v\cdot\nabla)v = -\nabla\Phi,
\]
\[
\partial_t\,S + \mathrm{div}(S\,v) = \Delta S,
\]
interpreted as potential relaxation, flow advection, and entropy
diffusion respectively. These equations specify lamphrodynamic
constraints on geometric evolution under collapse. Their precise form
may vary with the choice of background metric, constitutive relation,
or thermodynamic model; the above form is representative of the
structure assumed in the main text.

\section{Lamphrodynamic Evolution and Hamiltonian Structure}

Lamphrodynamics denotes the thermodynamically constrained evolution of
geometric configurations under collapse and merge in the presence of
the RSVP fields. A Hamiltonian functional capturing internal energy
and entropy release may be written
\[
\mathcal{H}[\Phi,v,S]
  = \int_P
    \Big(
       \tfrac{1}{2}\,|v|^2
       +\tfrac{1}{2}\,|\nabla\Phi|^2
       -S
    \Big)\, d\mathrm{vol},
\]
where the first two terms represent kinetic and potential energy
contributions, and the last term encodes entropy release. Collapse
reduces $\mathcal{H}$ by increasing boundary entropy while decreasing
internal potential.

In the main text, collapse is interpreted as a geometric operation
$D_{\mathrm{pop}}$ acting on spheres in $P$. In Hamiltonian form this
is realized as a vector field on function space,
\[
\dot{\Phi} = -\,\frac{\delta \mathcal{H}}{\delta \Phi}, \qquad
\dot{v} = -\,\frac{\delta \mathcal{H}}{\delta v}, \qquad
\dot{S} =  \frac{\delta \mathcal{H}}{\delta S},
\]
modulated by admissibility conditions reflecting boundary effects and
flow constraints. Collapse thus induces a Hamiltonian evolution
constrained by entropy production and flow compatibility. Under the
amplitwistor lift (Sections~\ref{sec:twistorsc}--\ref{sec:twistrsvp}),
this Hamiltonian vector field becomes a holomorphic differential
operator acting on the corresponding amplitwistor loci, producing a
coupled geometric--thermodynamic evolution in amplitwistor space.

\newpage
% -------------------------------------------------------------------
% Bibliography
% -------------------------------------------------------------------
\begin{thebibliography}{99}

\bibitem{Needham}
Needham, Tristan. \emph{Visual Complex Analysis}. Oxford University
Press, 1997.

\bibitem{PenroseTwistor}
Penrose, Roger. ``Twistor Algebra.'' \emph{Journal of Mathematical
Physics}, 8(2), 1967, 345--366.

\bibitem{WittenAmpl}
Witten, Edward. ``Perturbative Gauge Theory As A String Theory in
Twistor Space.'' \emph{Communications in Mathematical Physics}, 252
(2004): 189--258.

\bibitem{DomingosTensorLogic}
Domingos, Pedro. ``Tensor Logic: The Language of AI.''
\emph{arXiv:2510.12269} [cs.LG], 2025.

\bibitem{LambdaCalc}
Barendregt, H.~P. \emph{The Lambda Calculus: Its Syntax and
Semantics}. North--Holland, 1984.

\bibitem{GraphicalModels}
Koller, Daphne and Friedman, Nir. \emph{Probabilistic Graphical Models:
Principles and Techniques}. MIT Press, 2009.

\bibitem{DeepLearning}
Goodfellow, Ian; Bengio, Yoshua; Courville, Aaron. \emph{Deep
Learning}. MIT Press, 2016.

\end{thebibliography}

\end{document}


