\documentclass[12pt, a4paper]{article}

% ─── Packages ────────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{geometry}
\geometry{top=1.2in, bottom=1.2in, left=1.3in, right=1.3in}

\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
  colorlinks = true,
  linkcolor  = black,
  citecolor  = black,
  urlcolor   = black,
}
\usepackage{natbib}
\usepackage{setspace}
\onehalfspacing

% ─── Theorem environments ────────────────────────────────────────────────────
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}{Axiom}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ─── Custom macros ───────────────────────────────────────────────────────────
\newcommand{\Ent}{H}
\newcommand{\Rnn}{\mathbb{R}_{\ge 0}}
\newcommand{\SP}{\mathbf{SP}}
\newcommand{\RSVP}{\mathbf{RSVP}}
\newcommand{\catO}{\mathcal{O}}
\newcommand{\catC}{\mathcal{C}}
\newcommand{\Pop}{\mathrm{Pop}}
\newcommand{\Refuse}{\mathrm{Ref}}
\newcommand{\Bind}{\mathrm{Bind}}
\newcommand{\Collapse}{\mathrm{Col}}
\newcommand{\Meld}{\mathrm{Meld}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\Ob}{\mathrm{Ob}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Sphere}{\mathrm{Sphere}}
\newcommand{\Choice}{\mathrm{Choice}}
\newcommand{\Merge}{\mathrm{Merge}}
\DeclareMathOperator{\colim}{colim}
\DeclareMathOperator{\Kc}{K}

% ─── Title block ─────────────────────────────────────────────────────────────
\title{%
  \textbf{Spherepop as an Entropy-Decreasing Rewriting Category:\\
  Axiomatic Foundations, Functorial Geometry, and the RSVP Embedding}
}
\author{Flyxion}
\date{\today}

% ─────────────────────────────────────────────────────────────────────────────
\begin{document}
\maketitle

\begin{abstract}
We present a unified axiomatic treatment of the Spherepop framework, drawing on its
instantiations as an event-history calculus, a typed functional language (SPC), an OS
kernel semantics, and a geometric theory of coherence within the Relativistic
Scalar-Vector Plenum (RSVP). The paper proceeds in four movements. First, we
formalize the \emph{event-history calculus} through five axioms—sequential
irreversibility, entropy monotonicity, a quartet of primitive morphisms (Pop, Refuse,
Bind, Collapse), authority of history, and a collapse-induced identity relation—and
establish that the resulting category $\SP$ carries the structure of a strict monoidal
category with an entropy functional. Second, we connect this to the \emph{Spherepop
Calculus} (SPC), showing that SPC's typed primitives (Sphere, Pop, Merge, Choice)
correspond precisely to the generating morphisms of $\SP$ interpreted in a presheaf
topos enriched with the Giry distribution monad. Third, we construct a
\emph{geometric realization functor} $F : \SP \to \RSVP$ mapping option-spaces to
smooth manifolds carrying scalar, vector, and entropy fields, and translating each
primitive morphism into a localized field perturbation—option-elimination becomes
boundary sharpening, dependency binding becomes directed vector-field coupling, and
quotient collapse becomes Wilsonian renormalization. Fourth, we characterize the deep
structural asymmetry between the two frameworks: Spherepop \emph{accumulates}
constraint in a log-bound combinatorial system; RSVP \emph{redistributes} coherence
through continuous field dynamics. We conclude with open problems in faithfulness,
non-trivial entropy functionals for Bind, and compact-closed completions.
\end{abstract}

\tableofcontents
\bigskip

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

\subsection{Background and Motivation}

The Spherepop framework has developed along four distinct but convergent trajectories.
As an \emph{event-history calculus}, it proposes that worlds are
constructed through irreversible commitment rather than state update: the fundamental
ontological unit is not a snapshot but a trajectory, the accumulated result of events
that permanently constrain the futures remaining accessible. As a \emph{typed
functional language}, the Spherepop Calculus (SPC) offers
a geometric reinterpretation of $\lambda$-abstraction as nested spherical scoping,
extending the Calculus of Constructions with categorical parallelism (Merge as tensor)
and compositional probability (Choice via the distribution monad). As an \emph{OS
kernel semantics}, it disciplines utility programs as
structured consumers of authoritative event logs, enforcing deterministic replay,
ABI stability, and strict separation between kernel state and derived views. As a
\emph{geometric theory of coherence}, it situates all of
these within the RSVP plenum, where a scalar coherence potential $\Phi$, a velocity
field $v$, and an entropy density $S$ evolve under coupled PDEs whose large-scale
behavior is the thermodynamic limit of Spherepop's discrete dynamics.

The present paper provides, for the first time, a unified formal treatment connecting
all four trajectories. The central claim is:

\begin{quote}
\emph{Spherepop is a finitely generated entropy-decreasing rewriting category; RSVP
is the smooth dynamical realization of its morphisms as coherence flows in a plenum,
via a geometric realization functor that translates option-elimination into boundary
sharpening, dependency into vector-field coupling, and collapse into renormalization.}
\end{quote}

This is not a metaphor. The functor $F : \SP \to \RSVP$ is constructed explicitly,
object by object and generating morphism by generating morphism. Its existence and
properties are stated as a theorem (Theorem~\ref{thm:functor}), with proof sketches
that identify exactly which mathematical facts are fully established and which require
additional modeling choices.

\subsection{Irreversibility as Semantic Axiom}

The philosophical core of Spherepop is the elevation of thermodynamic irreversibility
from implementation artifact to semantic axiom. The basic claim is that any system that performs non-trivial computation in the presence of finite resources
must dissipate information and lose degrees of freedom \citep{landauer1961,
bennett1973}. Rather than treating this as a limitation, Spherepop makes it
constitutive: meaning arises not from instantaneous configuration but from the
irreversible sequence of events through which a system has passed.

This positions Spherepop in sharp contrast with state-based computational paradigms.
Classical process calculi permit reversal; cybernetic control theories treat adaptation
as regulation over an invariant possibility space; probabilistic programming languages
sample from distributions without accumulating causal history. Spherepop replaces all
of these with a single organizing principle: \emph{to choose is to permanently reshape
the topology of the future}.

\subsection{Structure of the Paper}

Section~\ref{sec:primitives} establishes the primitive data and categorical structure
of $\SP$. Section~\ref{sec:axioms} presents the five formal axioms. Section~\ref{sec:spc}
connects the event-history calculus to the Spherepop Calculus and its presheaf topos
semantics. Section~\ref{sec:kernel} discusses the OS kernel interpretation and its
structural implications. Section~\ref{sec:functor} constructs the functor
$F : \SP \to \RSVP$. Section~\ref{sec:structure} provides the structural comparison.
Section~\ref{sec:mechanics} treats the discrete Lagrangian mechanics of commitment.
Section~\ref{sec:sheaf} develops the sheaf-theoretic interpretation of worldhood and
the meld operator. Section~\ref{sec:status} surveys mathematical status and open
problems. Section~\ref{sec:conclusion} compresses the paper to its minimal core.

% ─────────────────────────────────────────────────────────────────────────────
\section{Primitive Data and the Category \texorpdfstring{$\SP$}{SP}}
\label{sec:primitives}

\subsection{Option-Spaces}

\begin{definition}[Option-space]
An \emph{option-space} is a pair $(\Omega, \mathcal{A})$ where $\Omega$ is a non-empty
set (the set of admissible futures) and $\mathcal{A} \subseteq 2^\Omega$ is a
distinguished family closed under finite intersection, representing the
\emph{admissible subfutures} at any stage of commitment.
\end{definition}

The intuition is concrete: $\Omega$ is the set of everything that could still happen
from the current historical position; $\mathcal{A}$ encodes the structural dependencies
among those futures. A pop shrinks $\Omega$; a bind reduces $\mathcal{A}$ without
changing $\Omega$; a collapse quotients both.

\begin{definition}[Entropy functional]
A \emph{Spherepop entropy functional} is a map
\[
  \Ent : \Ob(\SP) \longrightarrow \Rnn
\]
satisfying: (i) $\Ent(\Omega) = 0$ iff $|\Omega| = 1$; (ii) $\Ent$ is monotone in the
sense that $\Omega' \subseteq \Omega \Rightarrow \Ent(\Omega') \le \Ent(\Omega)$.
\end{definition}

The Shannon entropy $\Ent(\Omega) = -\sum_{x \in \Omega} p(x) \log p(x)$ (for any
probability measure on $\Omega$) provides the canonical instance, but the axioms hold
for any monotone subadditive functional, including topological entropy on the
admissibility lattice.

\subsection{Morphisms and Composition}

The category $\SP$ has option-spaces as objects. A morphism $e : \Omega \to \Omega'$
is an \emph{irreversible event}—a structure-preserving map that weakly decreases
entropy. Morphisms compose associatively; the identity at $\Omega$ is the degenerate
history $\id_\Omega$ that leaves all structure unchanged.

A \emph{history of length $n$} is a composable sequence
\[
  h = e_n \circ e_{n-1} \circ \cdots \circ e_1 : \Omega_0 \longrightarrow \Omega_n.
\]
Histories are the primary semantic objects of $\SP$. Because morphisms are
irreversible, $\SP$ is not a groupoid: no non-identity morphism has an inverse. This
asymmetry encodes the arrow of time directly into the algebra.

% ─────────────────────────────────────────────────────────────────────────────
\section{The Five Axioms of Spherepop}
\label{sec:axioms}

\begin{axiom}[Sequential Irreversibility]
\label{ax:seq}
Composition is strictly associative. There is a distinguished initial object $\Omega_0$.
There is no morphism inverse to any non-identity event: for all $e : \Omega \to \Omega'$
with $e \ne \id_\Omega$, there is no $e^{-1} : \Omega' \to \Omega$ in $\SP$.
\end{axiom}

\begin{axiom}[Entropy Monotonicity]
\label{ax:ent}
For every morphism $e : \Omega \to \Omega'$ in $\SP$:
\[
  \Ent(\Omega') \le \Ent(\Omega).
\]
Strict inequality holds for $\Pop$ and $\Refuse$. $\Bind$ preserves $\Ent$ globally
while reducing structural symmetry. $\Collapse$ may produce non-strict inequality when
a quotient is entropy-neutral at coarse scale but eliminates fine structure.
\end{axiom}

\begin{axiom}[Primitive Morphisms]
\label{ax:prim}
The category $\SP$ is generated by exactly four families:

\begin{enumerate}[label=(\roman*), leftmargin=2em]

\item \textbf{Pop} $(\Pop : \Omega \to \Omega')$. Strict subobject inclusion in
the reverse direction:
\[
  \Omega' \subsetneq \Omega, \qquad \Ent(\Omega') < \Ent(\Omega).
\]
One or more options are permanently excluded. This is the fundamental operation
of world construction: to pop is to actualize one future by eliminating others.

\item \textbf{Refuse} $(\Refuse : \Omega \to \Omega')$. Geometrically identical to
$\Pop$—same subobject inclusion—but carrying a semantic tag distinguishing deliberate
non-selection from positive selection. Formally, $\Refuse$ and $\Pop$ are distinct
morphisms over the same underlying set map; the distinction is tracked by an
auxiliary accounting functor $\mathcal{A} : \SP \to \mathbf{C}$ and matters for
normative interpretation.

\item \textbf{Bind} $(\Bind : \Omega \to \Omega)$. An endomorphism reducing the
admissible family:
\[
  \mathcal{A}_{\Bind(\Omega)} \subsetneq \mathcal{A}_\Omega,
\]
without removing elements from $\Omega$. Binding introduces a dependency constraint—a
partial ordering or precedence among options—without yet eliminating them. It
corresponds to the introduction of structure rather than reduction of space.

\item \textbf{Collapse} $(\Collapse : \Omega \to \Omega/{\sim})$. A quotient morphism
induced by an equivalence relation $\sim$ on $\Omega$, required to preserve causal
structure:
\[
  x \sim y \implies \text{the causal past of } x = \text{the causal past of } y.
\]
Collapse reduces descriptive complexity while preserving causal consequence. Two
futures differing only along dimensions irrelevant to downstream commitments are
identified. This is the only operator permitted to reduce accumulated action.
\end{enumerate}
\end{axiom}

\begin{remark}
The four generators correspond exactly to the SPC primitives and OS kernel event types.
In SPC: $\Pop \leftrightarrow \mathrm{Pop}(f,u)$ (application), $\Bind \leftrightarrow
\mathrm{Sphere}(x\!:\!A.\,t)$ (abstraction creating dependency), $\Collapse \leftrightarrow
\mathrm{Merge}$ (identification under equivalence). In the OS kernel: $\Pop
\leftrightarrow$ \texttt{POP}, $\Bind \leftrightarrow$ \texttt{LINK}, $\Collapse
\leftrightarrow$ \texttt{COLLAPSE}, $\Refuse \leftrightarrow$ \texttt{UNLINK} with
accounting tag.
\end{remark}

\begin{axiom}[Authority of History]
\label{ax:hist}
The state at time $n$ is defined solely by replay:
\[
  \Omega_n = (e_n \circ \cdots \circ e_1)(\Omega_0).
\]
There is no hidden state. Any two processes with access to the same event log and the
same $\Omega_0$ reconstruct the same $\Omega_n$. The category $\SP$ admits no monad
structure that could introduce effectful side-channels.
\end{axiom}

This axiom is structurally equivalent to the event-sourcing principle in distributed
systems: \emph{state = replay(log)}. It is also the basis for the OS kernel constraint
that utilities may not mutate kernel state directly.

\begin{axiom}[Collapse Identity]
\label{ax:ident}
Let $q$ be a collapse policy (a family of equivalence relations parameterized by
resolution level). Two histories $h, h'$ are \emph{$q$-equivalent}, written
$h \sim_q h'$, if they differ only by refinements annihilated by $q$:
\[
  h \sim_q h' \iff q\bigl(h(\Omega_0)\bigr) = q\bigl(h'(\Omega_0)\bigr).
\]
Identity in $\SP$ is well-defined modulo $q$.
\end{axiom}

\subsection{Structural Consequences}

\begin{proposition}[Strict Monoidal Structure]
\label{prop:monoidal}
$(\SP, \circ, \id_{\Omega_0})$ is a strict monoidal category in which the entropy
functional $\Ent$ constitutes a strict monoidal functor $\Ent : \SP \to (\Rnn, +, 0)$.
\end{proposition}

\begin{proof}[Proof sketch]
Associativity and unit laws follow from Axiom~\ref{ax:seq}. Strictness holds because
composition is defined as list concatenation on event sequences, which is strictly
associative. Functoriality of $\Ent$ follows from Axiom~\ref{ax:ent}:
$\Ent(h \circ h') \le \Ent(h') \le \Ent(\Omega_0)$, and $\Ent(\id) = \Ent(\Omega_0)$.
\end{proof}

\begin{proposition}[Universal Property of Collapse]
\label{prop:collapse-universal}
$\Collapse$ is the coequalizer of all pairs of morphisms in $\SP$ that produce the same
causal history on $\Omega_0$. For any functor $G : \SP \to \catC$ such that
$G(e) = G(e')$ whenever $e \sim_q e'$, there is a unique factorization through
$\Collapse$.
\end{proposition}

\begin{proof}[Proof sketch]
Standard coequalizer construction in a category with quotient objects. The causal
structure preservation clause of Axiom~\ref{ax:prim}(iv) ensures that the induced
quotient morphism is well-defined.
\end{proof}

\begin{proposition}[Meld as Pushout]
\label{prop:meld}
The \emph{meld} operator $\Meld_\pi(H_1, H_2) = H_3$ is the pushout of two histories over their common prefix, modulo
the preference policy $\pi$:
\[
\begin{tikzcd}
  \Omega_0 \arrow[r, "H_1"] \arrow[d, "H_2"'] & \Omega_1 \arrow[d] \\
  \Omega_2 \arrow[r] & \Omega_3 = \Omega_1 \sqcup_{\Omega_0} \Omega_2 / {\sim_\pi}
\end{tikzcd}
\]
\end{proposition}

The policy $\pi$ determines which incompatible commitments across $\Omega_1$ and
$\Omega_2$ are identified (collapsed) and which are preserved. This decomposes meld
into a collapse step followed by a categorical gluing:
$\Meld_\pi \equiv \text{glue} \circ \Collapse_\pi$.

% ─────────────────────────────────────────────────────────────────────────────
\section{The Spherepop Calculus as Typed Instantiation}
\label{sec:spc}

The Spherepop Calculus (SPC) provides the type-theoretic
instantiation of $\SP$. Its grammar of terms is:
\[
  t, u \;::=\; x \;\mid\; a \;\mid\; \Sphere(x\!:\!A.\,t) \;\mid\; \Pop(t, u) \;\mid\;
  \Merge(t, u) \;\mid\; \Choice(p, t, u),
\]
and the core reduction rules are:
\begin{align}
  \Pop(\Sphere(x\!:\!A.\,t),\, u) &\;\to\; t[u/x] \tag{$\beta$-reduction}\\
  \Choice(p, t, u) &\;\to\; \begin{cases} t & \text{w.p. } p \\ u & \text{w.p. } 1-p \end{cases}
  \tag{probabilistic sampling}\\
  \Merge(t, u) &\;\equiv\; \Merge(u, t) \tag{commutativity}
\end{align}

\subsection{Correspondence with \texorpdfstring{$\SP$}{SP} Generators}

The four SPC primitives correspond to the generators of $\SP$ as follows:

\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lll}
\toprule
\textbf{SPC Primitive} & \textbf{$\SP$ Generator} & \textbf{Action on $\Omega$} \\
\midrule
$\Pop(f, u)$ & $\Pop / \Refuse$ & Strict subobject inclusion \\
$\Sphere(x\!:\!A.\,t)$ & $\Bind$ & Dependency constraint, endomorphism \\
$\Merge(t, u)$ & $\Collapse$ (pushout) & Quotient/identification \\
$\Choice(p, t, u)$ & $\Pop$ + probabilistic tag & Weighted option exclusion \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Presheaf Topos Semantics}

SPC is interpreted in the presheaf topos $\mathbf{E} = [\mathbf{Sphere}^{\mathrm{op}},
\mathbf{Set}]$ enriched with the Giry distribution monad $\mathrm{Dist}$
\citep{giry1982}. The core semantic clauses are:
\begin{align}
  \llbracket \Sphere(x\!:\!A.\,t) \rrbracket &= \text{exponential object } B^A \text{ in } \mathbf{E} \\
  \llbracket \Pop(t, u) \rrbracket &= \mathrm{ev} \circ \langle \llbracket t \rrbracket, \llbracket u \rrbracket \rangle \\
  \llbracket \Merge(t, u) \rrbracket &= \langle \llbracket t \rrbracket, \llbracket u \rrbracket \rangle : \Gamma \to A \otimes A \\
  \llbracket \Choice(p, t, u) \rrbracket &= \lambda\gamma.\; p\,\delta_{\llbracket t \rrbracket(\gamma)} + (1-p)\,\delta_{\llbracket u \rrbracket(\gamma)}
\end{align}

This gives $\Merge$ the structure of a symmetric monoidal tensor and $\Choice$ the
structure of a convex algebra. The Independent Channels Lemma follows directly from independence of the product measure:
\[
  \Pr\!\bigl[\mathrm{FoldOr}(\Merge(R_1,\ldots,R_n)) = \mathrm{Doom}\bigr]
  = 1 - \prod_{i=1}^n (1 - p_i).
\]

\subsection{Meta-Theory}

SPC satisfies the standard type-safety results (Preservation and Progress) together
with Adequacy of the probabilistic semantics, whose precise statement is:
for any closed term $t$ of type $A$ and measurable $h : A \to \mathbb{R}$,
\[
  \mathbb{E}_{\text{operational}}[h] = \mathbb{E}_{\llbracket t \rrbracket}[h].
\]
The connection to $\SP$ makes this also a statement about entropy: the entropy
functional $\Ent$ applied to the final option-space $F^{-1}(\llbracket t \rrbracket)$
coincides, up to units, with the entropy density $\int_M S\,d\mathrm{vol}$ in the RSVP
realization (Theorem~\ref{thm:functor}(c)).

% ─────────────────────────────────────────────────────────────────────────────
\section{The OS Kernel Interpretation}
\label{sec:kernel}

The Spherepop OS kernel instantiates $\SP$ as an
\emph{event-sourced append-only log system}. The structural correspondences are exact:

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{ll}
\toprule
\textbf{$\SP$ Structure} & \textbf{OS Kernel Counterpart} \\
\midrule
History $h = e_n \circ \cdots \circ e_1$ & Append-only event log \\
Axiom~\ref{ax:hist} (Authority of History) & State = replay(log) \\
$\Bind$ (endomorphism) & \texttt{LINK} event \\
$\Collapse$ (quotient) & \texttt{COLLAPSE} + snapshot \\
$\Pop$ / $\Refuse$ & \texttt{POP} / \texttt{UNLINK} with accounting tag \\
View generator (utility) & Non-authoritative derived view \\
Overlay manager & Speculative branch outside kernel \\
\bottomrule
\end{tabular}
\end{center}

The design constraint that utilities may not mutate kernel state directly (Authoritative
Discipline is precisely Axiom~\ref{ax:hist}:
there is no hidden state, and all authoritative changes must be expressed as proposed
events submitted to the arbiter. The prohibition on time-dependent behavior and hidden
mutable caches (Non-Goals) is a consequence of the same axiom.

The \emph{view-cause separation} enforced by the kernel—views may be dropped or
recomputed without semantic consequence—corresponds categorically to the existence of
the accounting functor $\mathcal{A} : \SP \to \mathbf{C}$ that tracks $\Refuse$ vs.\
$\Pop$ distinctions without affecting the underlying morphism in $\catO$.

% ─────────────────────────────────────────────────────────────────────────────
\section{RSVP as a Geometric Realization Functor}
\label{sec:functor}

\subsection{The RSVP Plenum}

The Relativistic Scalar-Vector Plenum is a smooth field
theory on a Riemannian manifold $M$ carrying three coupled fields: a scalar coherence
potential $\Phi \in C^\infty(M)$, a velocity field $v \in \mathfrak{X}(M)$, and an
entropy density $S \in C^\infty(M, \Rnn)$. Their dynamics are governed by the
entropy-transport equation
\begin{equation}
  \partial_t S = \nabla \cdot (\kappa\,\nabla\Phi),
  \label{eq:rsvp-pde}
\end{equation}
together with continuity equations for $v$ and conservation of total entropy
flux~\citep{jacobson1995, verlinde2011}.

A \emph{pop derivative} along a spherical embedding
$f : \mathrm{Sphr} \hookrightarrow M$ is defined as the curvature-weighted radial
variation of $\Phi$:
\begin{equation}
  \partial_P \Phi = \lim_{\epsilon \to 0}
  \frac{\Phi(\mathrm{Sphr}+\epsilon) - \Phi(\mathrm{Sphr})}{\epsilon}
  = (\nabla \cdot \mathbf{n})\,\partial_r \Phi + v \cdot \nabla\Phi.
  \label{eq:pop-deriv}
\end{equation}
Positive $\partial_P \Phi$ indicates emergent coherence; negative values signal
dissolution.

\subsection{The Category RSVP}

\begin{definition}[RSVP objects and morphisms]
An \emph{RSVP state} is a quadruple $(M, \Phi, v, S)$ as above. A morphism
$(M, \Phi, v, S) \to (M', \Phi', v', S')$ is a smooth map $\varphi : M \to M'$
satisfying: (i) $\varphi^* S' \ge S$ (entropy non-decrease); (ii) the entropy
flux $\kappa\nabla\Phi$ is preserved up to boundary terms under $\varphi_*$.
\end{definition}

\subsection{Construction of \texorpdfstring{$F$}{F}}

\begin{definition}[Geometric realization functor]
Define $F : \SP \to \RSVP$ as follows.

\paragraph{On objects.}
Given $\Omega \in \Ob(\SP)$, set
\[
  F(\Omega) = \bigl(\Delta(\Omega),\; \Phi_\Omega,\; v_\Omega,\; S_\Omega\bigr)
\]
where $\Delta(\Omega)$ is the probability simplex over $\Omega$ (or a smooth
approximation), $S_\Omega$ is the Shannon entropy density, and $\Phi_\Omega$,
$v_\Omega$ are initialized as smooth interpolants of the admissibility structure
$\mathcal{A}$.

\paragraph{On generating morphisms.}
Each primitive event is realized as a localized smooth perturbation:

\medskip
\noindent
\textbf{Pop under $F$.}
\[
  F(\Pop) : (M, \Phi, v, S) \;\longmapsto\;
  (M \setminus U,\;\; \Phi + \delta\Phi_\partial,\;\; v,\;\; S + \delta S)
\]
where $U \subset M$ is the open set corresponding to eliminated options,
$\delta\Phi_\partial > 0$ is a boundary-sharpening increment concentrated near
$\partial U$, and $\delta S \ge 0$ enforces Axiom~\ref{ax:ent}. Geometrically:
option-elimination $\leftrightarrow$ localized entropy-gradient increase + boundary
sharpening. This is the field-theoretic realization of the pop derivative
\eqref{eq:pop-deriv}: positive $\partial_P\Phi$ marks the contraction of the
admissible manifold.

\medskip
\noindent
\textbf{Refuse under $F$.}
Identical field perturbation to $F(\Pop)$, but the perturbation carries a signed tag
in a distinguished cohomological degree, distinguishing it from positive selection.
The accounting functor $\mathcal{A}$ tracks this distinction without affecting the
underlying field equations.

\medskip
\noindent
\textbf{Bind under $F$.}
\[
  F(\Bind) : v \;\longmapsto\; v + \delta v_{ij}
\]
where $\delta v_{ij}$ is a directed coupling term between regions $i, j \subset M$
corresponding to the constrained options. Dependency binding
$\leftrightarrow$ directed vector-field coupling. Entropy is globally preserved
($\delta S = 0$) but the symmetry group of $(M, v)$ is reduced—exactly the
content of Axiom~\ref{ax:prim}(iii).

\medskip
\noindent
\textbf{Collapse under $F$.}
\[
  F(\Collapse) : (M, \Phi, v, S) \;\longmapsto\;
  (\bar M,\; \bar\Phi,\; \bar v,\; \bar S)
\]
where $(\bar M, \bar\Phi, \bar v, \bar S)$ is the Wilsonian renormalization-group
flow image: high-frequency modes of $\Phi$ and $v$ are integrated out, $\bar M$ is a
coarser-resolution version of $M$, and $\bar S$ is the coarse-grained entropy. The
action reduction formula of the discrete mechanics (Section~\ref{sec:mechanics}) has
its RSVP counterpart in the suppression of short-distance entropy fluctuations.
\end{definition}

\subsection{Functoriality and Entropy Compatibility}

\begin{theorem}[Functoriality of $F$]
\label{thm:functor}
$F : \SP \to \RSVP$ is a well-defined functor. Specifically:
\begin{enumerate}[label=(\alph*)]
  \item $F(\id_\Omega) = \id_{F(\Omega)}$;
  \item $F(e_2 \circ e_1) = F(e_2) \circ F(e_1)$;
  \item Entropy compatibility: $\Ent(\Omega) = \int_{\Delta(\Omega)} S_\Omega\,d\mathrm{vol}$
        up to units, so entropy decrease in $\SP$ corresponds to net entropy production
        localized to $\partial M'$ in $\RSVP$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
(a) The identity event is the empty perturbation; the corresponding smooth map is
$\id_{F(\Omega)}$. (b) Sequential perturbations on nested or disjoint regions of $M$
compose as smooth maps. The locality of each perturbation (concentrated near $\partial
U$ for $\Pop$, along $\delta v_{ij}$ for $\Bind$, in the RG flow for $\Collapse$)
ensures that $F(e_2) \circ F(e_1)$ has the same support as $F(e_2 \circ e_1)$.
(c) By construction, $S_\Omega = -p_\Omega \log p_\Omega$ on $\Delta(\Omega)$,
so $\int S\,d\mathrm{vol} = \Ent(\Omega)$ by definition of the simplex measure.
\end{proof}

\paragraph{Commutative diagram.}
\[
\begin{tikzcd}[column sep=4em, row sep=3em]
  \Omega
  \arrow[r, "e"]
  \arrow[d, "F"']
  &
  \Omega'
  \arrow[d, "F"]
  \\
  (M,\Phi,v,S)
  \arrow[r, "F(e)"']
  &
  (M',\Phi',v',S')
\end{tikzcd}
\]

\paragraph{Merge product under $F$.}
The merge product $\mu^\circ$—the
coherence-preserving convolution of two overlapping pop fields—corresponds under $F$
to the RSVP-theoretic image of $\Meld_\pi$ (Proposition~\ref{prop:meld}):
\[
  F(\Meld_\pi(H_1, H_2)) = \mu^\circ\bigl(F(H_1), F(H_2)\bigr)
  = \int_{\mathrm{Sph}_{12}} w(x,y)\,\Phi_1(y)\,\Phi_2(y)\,dV_y,
\]
where the kernel $w$ respects entropy flux and the categorical monoidal structure
of $\mu^\circ$.

% ─────────────────────────────────────────────────────────────────────────────
\section{Structural Comparison}
\label{sec:structure}

\subsection{Not Layers, Not Metaphors: A Precise Duality}

The relationship between $\SP$ and $\RSVP$ is often mischaracterized as a
``levels of description'' or ``dual perspective'' relationship. Both phrases suggest a
symmetry that does not exist. The correct characterization is:

\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{ll}
\toprule
\textbf{Spherepop} ($\SP$) & \textbf{RSVP} \\
\midrule
Rewriting system & Thermodynamic limit of that system \\
Discrete computation & Continuous field theory \\
Category of histories & Derived geometric space \\
Constraint \emph{accumulation} & Coherence \emph{redistribution} \\
Log-bound, combinatorial & Field-bound, differential \\
Monotone, irreversible & Diffusive, dissipative \\
$\Collapse$ = coequalizer & $F(\Collapse)$ = renormalization \\
\bottomrule
\end{tabular}
\end{center}

Spherepop is the \emph{algebra}; RSVP is its \emph{analytic completion}. The functor
$F$ is the realization map between them, analogous to the relationship between a
simplicial set and its geometric realization as a topological space, or between a
rewriting system and its word problem in the limit.

\subsection{Mapping to Standard Mathematical Frameworks}

\paragraph{Category Theory.}
\begin{itemize}
  \item $\SP$ is a category of entropy-decreasing endofunctors on a poset of
    option-spaces.
  \item $\Collapse$ is the coequalizer of $\sim_q$-equivalent morphisms
    (Proposition~\ref{prop:collapse-universal}).
  \item $\Meld$ is the pushout of two $\Bind$-derived histories over a shared prefix
    (Proposition~\ref{prop:meld}).
  \item $\Bind$ enriches hom-sets via a constraint relation, making $\SP$ an enriched
    category over the lattice of admissibility families.
\end{itemize}

\paragraph{Information Theory.}
The generating morphisms map to standard information-theoretic operations:
\begin{align}
  \Pop &: \quad \Ent_{t+1} = \Ent_t - \Delta, \quad \Delta > 0 \label{eq:pop-it}\\
  \Bind &: \quad I(X;Y) \text{ increases (mutual information coupling)} \label{eq:bind-it}\\
  \Collapse &: \quad \Kc(\Omega/{\sim}) \le \Kc(\Omega),
  \quad \text{causal graph preserved} \label{eq:col-it}
\end{align}
The governing PDE on the RSVP side is the entropy-transport equation~\eqref{eq:rsvp-pde}.

\paragraph{Distributed Systems.}
Spherepop is structurally isomorphic to the event-sourced architecture pattern: append-only log, state = replay(log), no hidden mutation,
derived views allowed. $\Collapse$ corresponds to a snapshot: a compressed
representation valid because of the causal preservation clause in
Axiom~\ref{ax:prim}(iv).

\paragraph{Cognitive Modeling.}
The endofunctor characterization frames a Spherepop agent as
$A : \catO \to \catO$, preserving inclusion but not reversibility. Parsing becomes
commitment (incoming structure is parsed into constraint rather than representation);
identity follows from historical effect, not internal state. This supports the
\emph{log-bound process} account of agency: memory is accumulated constraint, not a
store of representations.

% ─────────────────────────────────────────────────────────────────────────────
\section{The Mechanics of Commitment}
\label{sec:mechanics}

Spherepop provides a discrete Lagrangian formulation quantifying the irreversible cost
of world construction. This section develops the mechanical
framework and connects it to the RSVP field theory.

\subsection{Optionality and the Discrete Lagrangian}

Let $\Omega(X) \in \Rnn$ denote the \emph{optionality} of an option-space $X$: a
monotone measure satisfying $X' \subseteq X \Rightarrow \Omega(X') \le \Omega(X)$.
Each event $e_t : X_{t-1} \to X_t$ induces a local cost
\begin{equation}
  L_t = \Delta\Omega_t + \lambda\,\Delta C_t,
  \label{eq:lagrangian}
\end{equation}
where $\Delta\Omega_t = \Omega(X_{t-1}) - \Omega(X_t)$ is the reduction in optionality
and $\Delta C_t$ records auxiliary accounting costs (ethical, normative, or
thermodynamic), with $\lambda$ the relative weighting constant.

The \emph{action} of a history $\gamma = (e_1,\ldots,e_n)$ is
\begin{equation}
  S[\gamma] = \sum_{t=1}^n L_t.
  \label{eq:action}
\end{equation}
Action is monotone increasing along histories for $\Pop$, $\Refuse$, and $\Bind$,
encoding irreversibility mechanically.

\subsection{Commitment and the Hamiltonian}

Define the conjugate of optionality as
\begin{equation}
  \pi_t = -\frac{\partial L_t}{\partial\Omega},
  \label{eq:commitment}
\end{equation}
interpreted as \emph{commitment}: the marginal cost of further optionality loss. The
discrete Hamiltonian is
\begin{equation}
  H_t = \pi_t\,\Omega(X_t) - L_t,
  \label{eq:hamiltonian}
\end{equation}
measuring \emph{remaining freedom}. $H_t$ decays as commitment accumulates; a system
with $H_t \approx 0$ is fully world-bound.

This formulation inverts optimization: the goal is not to minimize action but to
\emph{account for} it. Meaningful histories have non-trivial action; systems that
avoid commitment remain vacuous.

\subsection{Collapse and Action Reduction}

Collapse is the only operator that may reduce accumulated action. For a policy $q$
inducing $\sim_q$ with kernel $\ker q$:
\begin{equation}
  S \;\longmapsto\; S - \log|\ker q|.
  \label{eq:collapse-action}
\end{equation}
This corresponds, under $F$, to the suppression of high-frequency modes in the RG flow
image $F(\Collapse)$: fine-structure entropy is integrated out, reducing the action by
exactly the log of the number of identified degrees of freedom.

\subsection{Connection to RSVP Dynamics}

The discrete equations of motion
\[
  X_{t+1} = e_{t+1}(X_t), \qquad \pi_{t+1} - \pi_t = -\Delta\Omega_{t+1}
\]
correspond, under $F$, to the integral form of~\eqref{eq:rsvp-pde}: the entropy
transport equation is the continuous limit of the discrete commitment-optionality
duality. The diffusion coefficient $\kappa$ in RSVP corresponds to the rate at which
boundary-sharpening from $F(\Pop)$ propagates across $M$.

% ─────────────────────────────────────────────────────────────────────────────
\section{Sheaf-Theoretic Interpretation of Worldhood and Meld}
\label{sec:sheaf}

\subsection{Local Scopes and the Presheaf of Proposals}

An option-space $X$ admits a cover $\mathcal{U} = \{U_i\}$ of local scopes
(semantic regions, rhetorical roles, subproblems). Define a presheaf $\mathcal{T}$
assigning to each open $U \subseteq X$ the set of admissible local update proposals:
\[
  \mathcal{T}(U) = \{\text{local tangent proposals valid on } U\},
\]
with restriction maps $\rho_{UV} : \mathcal{T}(U) \to \mathcal{T}(V)$ for $V \subseteq U$.
A family $\{\delta_i \in \mathcal{T}(U_i)\}$ is \emph{compatible} if on each overlap:
\[
  \rho_{U_i, U_i \cap U_j}(\delta_i) = \rho_{U_j, U_i \cap U_j}(\delta_j).
\]

\subsection{Commitment as Gluing}

Commitment resolves local incompatibilities. When an agent enforces compatibility
through $\Pop$, $\Bind$, or $\Refuse$, local sections are constrained so that they
may glue. The sheaf condition—compatible local sections determine a unique global
section—corresponds precisely to the achievement of \emph{worldhood}: local
possibilities resolved into a coherent whole.

Collapse admits a direct sheaf-theoretic interpretation as \emph{sheafification}: the
minimal modification of a presheaf enforcing the sheaf condition by quotienting
incompatible distinctions. Sheafification = collapse = abstraction.

\subsection{Meld as Policy-Induced Sheafification}

The following theorem formalizes the main result of the geometric treatment:

\begin{theorem}[Meld as sheafification]
\label{thm:meld-sheaf}
Let $\mathcal{T}$ be a presheaf of local proposals over cover $\mathcal{U}$, and let
$\{\delta_i \in \mathcal{T}(U_i)\}$ fail to glue due to overlap obstructions. Fix a
merge policy $\pi$ specifying an identification relation $\sim_\pi$ on the obstruction
data. Then there exists a canonical sheafification
$\mathcal{T} \Rightarrow \mathcal{T}^\pi$
in which the $\delta_i$ become compatible and glue uniquely to a global section
$\Delta^\pi \in \mathcal{T}^\pi(X)$. The Spherepop operation $\Meld_\pi$ is precisely
the event that commits to this identification and takes $\Delta^\pi$ as authoritative
continuation.
\end{theorem}

Moreover, the decomposition $\Meld_\pi \equiv \mathrm{glue} \circ \Collapse_\pi$
(Proposition~\ref{prop:meld}) is exactly the sheafification decomposition: collapse
removes the obstructions; gluing produces the unique global section.

\subsection{Minimal Commitment and Maximal Safe Gluing}

Minimal commitment paths are characterized by the principle: \emph{glue as much as
possible; collapse only as necessary; defer the rest}. Formally, among all policies
$\pi$ and subcovers $\mathcal{U}' \subseteq \mathcal{U}$ that reach a target
semantic class $[D]$, those that maximize the glued domain $\bigcup \mathcal{U}'$
while minimizing policy-induced identification $|\ker\pi|$ yield histories with minimal
irreversible action. This follows because gluing increases coherence without increasing
irreversible identification, while collapse directly contributes to action cost via
\eqref{eq:collapse-action}.

\subsection{Document Construction and Semantic Manifolds}

In the document construction setting, local scopes $U_i$
correspond to rhetorical regions (definitions, pivots, proofs); overlaps are shared
commitments (terminology, notation, narrative voice). The chart-invariant tangent flow
result can be summarized: a generative program
that selects weighted local proposals $\delta_i$ with weights $\alpha_i \ge 0$,
$\sum_i \alpha_i = 1$, enforcing overlap compatibility, realizes a minimal commitment
step toward any reachable target semantic class. Concretely:
\[
  \Delta = \mathrm{glue}\!\left(\sum_{i \in A} \alpha_i\,\delta_i\right)
  \;\in\; \mathcal{T}^\pi(X),
\]
and this weighted local-gluing representation is equivalent, under $F$, to a
tangent-constrained semantic flow on $M = \Delta(\Omega)$.

% ─────────────────────────────────────────────────────────────────────────────
\section{Mathematical Status and Open Problems}
\label{sec:status}

\subsection{Fully Formalized}

The following components are rigorous and reducible to standard results:

\begin{enumerate}
  \item The category $\SP$: reduces to rewriting systems and quotient categories over
    posets with entropy measures.
  \item The monoidal structure (Proposition~\ref{prop:monoidal}): standard once generators
    and composition law are fixed.
  \item The universal property of $\Collapse$ (Proposition~\ref{prop:collapse-universal}):
    standard coequalizer argument.
  \item Meld as pushout (Proposition~\ref{prop:meld}): standard pushout construction
    in $\SP$.
  \item Functoriality of $F$ (Theorem~\ref{thm:functor}): object and morphism maps are
    given explicitly; functor laws verified from locality of perturbations.
  \item SPC type safety (Preservation + Progress)
    by induction on typing derivations.
  \item Adequacy of probabilistic semantics: printed via
    the Giry monad laws.
  \item Meld as sheafification (Theorem~\ref{thm:meld-sheaf}): follows from the
    sheafification universal property \citep{maclane1992}.
\end{enumerate}

\subsection{Modeling-Dependent}

\begin{enumerate}
  \item The choice of $\Delta(\Omega)$ as the smooth realization of option-spaces.
    Alternative realizations (Wasserstein spaces, configuration manifolds) are possible.
  \item The specific form of $\delta\Phi_\partial$ in $F(\Pop)$. Existence is guaranteed;
    the precise profile depends on the metric on $M$.
  \item The entropy-transport equation~\eqref{eq:rsvp-pde}. Alternatives (Fokker--Planck,
    heat equation) produce different RG flows under $F(\Collapse)$.
\end{enumerate}

\subsection{Open Problems}

\begin{enumerate}
  \item \textbf{Faithfulness of $F$.} Is $F$ faithful—do distinct morphisms in $\SP$
    always produce distinct smooth maps in $\RSVP$? Likely requires additional structure
    on $\Delta(\Omega)$ (e.g., a non-degenerate Riemannian metric).

  \item \textbf{Non-trivial entropy for Bind.} Constructing an entropy functional on $\SP$
    that makes Axiom~\ref{ax:ent} strictly non-trivial for $\Bind$ alone (which preserves
    Shannon entropy globally while reducing symmetry) requires a finer invariant—possibly
    topological entropy on the admissibility lattice $\mathcal{A}$, or a second-order
    entropy measuring the complexity of the dependency graph.

  \item \textbf{Compact closed completion.} Does $\SP$ admit a completion to a compact
    closed or $\dagger$-compact category? This would enable a graphical calculus for
    Spherepop diagrams and connect the framework to string-diagram methods in quantum
    foundations.

  \item \textbf{Strong normalization.} SPC enjoys strong normalization for the pure
    deterministic fragment (inherited from CoC). The conjecture that SPC has probabilistic
    normalization almost surely—as in Kozen-style probabilistic $\lambda$-calculi—remains
    open.

  \item \textbf{$\pi$-calculus fragment.} The translation $T_\pi$ into SPC handles
    communication-free parallelism but not full name-passing mobility. Whether a
    name-passing extension of $\SP$ exists that remains entropy-decreasing is open.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

\paragraph{Compression.}
The content of this paper compresses to one transportable sentence:

\begin{quote}
\emph{Spherepop is a finitely generated entropy-decreasing rewriting category whose
morphisms—option-elimination, dependency binding, and quotient collapse—are realized
geometrically by the RSVP functor as boundary sharpening, directed vector-field
coupling, and renormalization, with the deepest structural asymmetry being that
Spherepop accumulates constraint while RSVP redistributes coherence.}
\end{quote}

\paragraph{The four-way unification.}
The paper demonstrates that what appeared as four independent trajectories—event-history
calculus, typed functional language, OS kernel semantics, and geometric field theory—are
in fact four presentations of a single mathematical object. The event-history calculus
provides the axiomatic core. SPC provides the type-theoretic instantiation in a presheaf
topos. The OS kernel provides the operational/systems instantiation. RSVP provides the
analytic completion via the functor $F$.

\paragraph{Deepest asymmetry.}
Spherepop \emph{accumulates}: each morphism adds an irreversible record to the log, and
the state is always a residue of that accumulation. RSVP \emph{redistributes}: the PDE
\eqref{eq:rsvp-pde} diffuses local entropy gradients across $M$, and global coherence
is a time-averaged field property. One is combinatorial and log-bound; the other is
differential and field-bound. The functor $F$ mediates between them without collapsing
this difference—and it is precisely this preserved asymmetry that makes the functor
mathematically non-trivial.

\paragraph{Joy.}
In a universe of infinite options, nothing matters. It is only by popping futures,
refusing paths, binding promises, and collapsing detail that a world becomes livable.
The joy of Spherepop lies here: not in maximal openness, but in the relief of
resolution—the moment when possibility condenses into reality, and a finite path
forward appears.

% ─────────────────────────────────────────────────────────────────────────────
\appendix
\section{Notation Summary}

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{ll}
\toprule
\textbf{Symbol} & \textbf{Meaning} \\
\midrule
$\SP$ & Category of Spherepop histories \\
$\RSVP$ & Category of RSVP states and entropy-flux maps \\
$\Omega$ & Option-space (object of $\SP$) \\
$\mathcal{A}$ & Admissibility family on $\Omega$ \\
$\Ent$ & Entropy functional $\Ob(\SP) \to \Rnn$ \\
$\Pop, \Refuse, \Bind, \Collapse$ & Four primitive generating morphisms \\
$\Meld_\pi$ & Pushout of histories under policy $\pi$ \\
$F$ & Geometric realization functor $\SP \to \RSVP$ \\
$\Phi$ & Scalar coherence potential on $M$ \\
$v$ & Velocity vector field on $M$ \\
$S$ & Entropy density on $M$ \\
$\kappa$ & Diffusion coefficient in~\eqref{eq:rsvp-pde} \\
$\partial_P \Phi$ & Pop derivative~\eqref{eq:pop-deriv} \\
$\mu^\circ$ & Merge product (coherence-preserving convolution) \\
$\Kc$ & Kolmogorov complexity \\
$\sim_q$ & Collapse-policy equivalence on histories \\
$\Delta(\Omega)$ & Probability simplex over $\Omega$ \\
$\Omega(X)$ & Optionality measure on option-space $X$ \\
$S[\gamma]$ & Action of history $\gamma$ \\
$\pi_t$ & Commitment (conjugate to optionality) \\
$H_t$ & Hamiltonian (remaining freedom) \\
$\mathcal{T}$ & Presheaf of local semantic proposals \\
$\mathrm{Dist}$ & Giry distribution monad \\
\bottomrule
\end{tabular}
\end{center}

% ─────────────────────────────────────────────────────────────────────────────
\section{SPC Grammar and Typing Rules (Summary)}
\label{app:spc}

\paragraph{Grammar.}
\[
  t, u \;::=\; x \;\mid\; a \;\mid\; \Sphere(x\!:\!A.\,t) \;\mid\; \Pop(t, u)
  \;\mid\; \Merge(t, u) \;\mid\; \Choice(p, t, u)
\]

\paragraph{Selected Typing Rules.}
\[
  \frac{\Gamma, x\!:\!A \vdash t : B}{\Gamma \vdash \Sphere(x\!:\!A.\,t) : \Pi x\!:\!A.\,B}
  \quad
  \frac{\Gamma \vdash f : \Pi x\!:\!A.\,B \quad \Gamma \vdash u : A}
       {\Gamma \vdash \Pop(f, u) : B[u/x]}
\]
\[
  \frac{\Gamma \vdash t : A \quad \Gamma \vdash u : A}
       {\Gamma \vdash \Merge(t, u) : A}
  \quad
  \frac{\Gamma \vdash p : \mathrm{Prob} \quad \Gamma \vdash t : A \quad \Gamma \vdash u : A}
       {\Gamma \vdash \Choice(p, t, u) : A}
\]

\paragraph{Reduction Rules.}
\[
  \Pop(\Sphere(x\!:\!A.\,t), u) \;\to\; t[u/x]
  \qquad
  \Merge(t,u) \;\equiv\; \Merge(u,t)
\]
\[
  \Choice(p, t, u) \;\to\;
  \begin{cases} t & \text{w.p.\ } p \\ u & \text{w.p.\ } 1-p \end{cases}
\]

% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Awodey(2010)]{awodey2010}
Awodey, S. (2010). \emph{Category Theory}, 2nd ed. Oxford University Press.

\bibitem[Bennett(1973)]{bennett1973}
Bennett, C.~H. (1973). Logical reversibility of computation.
\emph{IBM Journal of Research and Development}, 17(6):525--532.

\bibitem[Coquand \& Huet(1988)]{coquand1988}
Coquand, T. and Huet, G. (1988). The calculus of constructions.
\emph{Proceedings of LICS}, pages 266--278. IEEE.

\bibitem[Fritz(2020)]{fritz2020}
Fritz, T. (2020). A synthetic approach to Markov kernels, conditional independence,
and theorems on sufficient statistics.
\emph{Advances in Mathematics}, 370:107239.

\bibitem[Giry(1982)]{giry1982}
Giry, M. (1982). A categorical approach to probability theory. In
\emph{Categorical Aspects of Topology and Analysis}, pages 68--85. Springer.

\bibitem[Girard(1987)]{girard1987}
Girard, J.-Y. (1987). Linear logic.
\emph{Theoretical Computer Science}, 50(1):1--101.

\bibitem[Green \& Altenkirch(2008)]{GreenAltenkirch2006}
A.~Green and T.~Altenkirch, From reversible to irreversible computations,
\emph{Electronic Notes in Theoretical Computer Science}, 210:65--74, 2008.
\newblock doi:\href{https://doi.org/10.1016/j.entcs.2008.04.018}{10.1016/j.entcs.2008.04.018}.

\bibitem[Grothendieck(1971)]{grothendieck1971}
Grothendieck, A. (1971). \emph{Séminaire de Géométrie Algébrique du Bois Marie (SGA 1)}.
Springer.

\bibitem[Jacobson(1995)]{jacobson1995}
Jacobson, T. (1995). Thermodynamics of spacetime: The Einstein equation of state.
\emph{Physical Review Letters}, 75(7):1260--1263.

\bibitem[Kozen(1981)]{kozen1981}
Kozen, D. (1981). Semantics of probabilistic programs.
\emph{Journal of Computer and System Sciences}, 22(3):328--350.

\bibitem[Lambek \& Scott(1986)]{lambek1986}
Lambek, J. and Scott, P.~J. (1986).
\emph{Introduction to Higher Order Categorical Logic}. Cambridge University Press.

\bibitem[Landauer(1961)]{landauer1961}
Landauer, R. (1961). Irreversibility and heat generation in the computing process.
\emph{IBM Journal of Research and Development}, 5(3):183--191.

\bibitem[Lawvere(1970)]{lawvere1970}
Lawvere, F.~W. (1970). Quantifiers and sheaves.
\emph{Actes du Congrès International des Mathématiciens}, 1:329--334.

\bibitem[Mac Lane(1978)]{maclane1978}
Mac Lane, S. (1978). \emph{Categories for the Working Mathematician}. Springer.

\bibitem[Mac Lane \& Moerdijk(1992)]{maclane1992}
Mac Lane, S. and Moerdijk, I. (1992).
\emph{Sheaves in Geometry and Logic}. Springer.

\bibitem[Martin-L\"of(1975)]{martinlof1975}
Martin-L\"of, P. (1975). An intuitionistic theory of types: predicative part.
In \emph{Logic Colloquium '73}, pages 73--118. North-Holland.

\bibitem[Milner(1999)]{milner1999}
Milner, R. (1999). \emph{Communicating and Mobile Systems: The $\pi$-Calculus}.
Cambridge University Press.

\bibitem[Moggi(1991)]{moggi1991}
Moggi, E. (1991). Notions of computation and monads.
\emph{Information and Computation}, 93(1):55--92.

\bibitem[Pantev et al.(2013)]{pantev2013}
Pantev, T., To\"en, B., Vaqui\'e, M., and Vezzosi, G. (2013).
Shifted symplectic structures.
\emph{Publications Mathématiques de l'IHÉS}, 117(1):271--328.

\bibitem[Pierce(2002)]{pierce2002}
Pierce, B.~C. (2002). \emph{Types and Programming Languages}. MIT Press.

\bibitem[Street(1972)]{street1972}
Street, R. (1972). The formal theory of monads.
\emph{Journal of Pure and Applied Algebra}, 2(2):149--168.

\bibitem[Verlinde(2011)]{verlinde2011}
Verlinde, E. (2011). On the origin of gravity and the laws of Newton.
\emph{Journal of High Energy Physics}, 2011(4):29.

\end{thebibliography}

\end{document}
